---
title: "A quantitative synthesis of wetland methane production and consumption rates with regression tree driver analysis—full reproducible code"
author: "Satya Kent"
date: "2025-3-14"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Abbreviations:
MOx = aerobic methane oxidation
AOM = anaerobic methane oxidation

Load data and clean for modeling
```{r}

#Load data 
library(tidyverse)
library(dplyr)

flux_rates=read_csv("flux_rates_meta-analysis_final.csv") 

# Specify the order of x-axis categories
desired_order2 <- c("Methanogenesis", "Aerobic methane oxidation", "Anaerobic methane oxidation")

# Convert "Pathway" column to factor with desired order
flux_rates$Pathway <- factor(flux_rates$Pathway, levels = desired_order2)

#create filtered dataframe with all the environmental variables I care about.
flux_rates_filtered <- flux_rates[, c("Pathway", "Mean_rate", "Wetland_type", "Inc_temp", "Inc_length_days","Depth_representative", "Salinity_ppt", "Salinity_cat", "dom_veg", "method","subpathway")]

#keep NAs in categorical values

# remove the continuous salinity column
flux_rates_filtered <- flux_rates_filtered %>%
  dplyr::select(-Salinity_cat)

#re-organize datasets for modeling of each pathway separately.

#hot encode categorical values to integers (machine learning models generally perform better when hot encoding categorical predictors, plus xgboost requires this)
#Save the factor levels for categorical variables before encoding (we'll need later to plot the PDP's for categorical predictors)

#MGEN
MGEN_filtered <- flux_rates_filtered[flux_rates_filtered$Pathway == "Methanogenesis", ]
MGEN_filtered$Pathway <- NULL
names(MGEN_filtered) <- c("Mean_rate", "Wetland_type", "Incubation_temperature", "Incubation_length", "Depth", "Salinity", "Dominant_vegetation", "Method", "Subpathway")

wetland_type_levels_MGEN <- levels(factor(MGEN_filtered$Wetland_type))
dom_veg_levels_MGEN <- levels(factor(MGEN_filtered$Dominant_vegetation))
method_levels_MGEN <- levels(factor(MGEN_filtered$Method))
subpathway_levels_MGEN <- levels(factor(MGEN_filtered$Subpathway))

MGEN_encoded <- MGEN_filtered %>%
  mutate(
    Wetland_type = as.integer(factor(Wetland_type)),
    Dominant_vegetation = as.integer(factor(Dominant_vegetation)),
    Method = as.integer(factor(Method)),
    Subpathway = as.integer(factor(Subpathway))
  )

#MOX
MOX_filtered=flux_rates_filtered[flux_rates_filtered$Pathway == "Aerobic methane oxidation", ]
MOX_filtered$Pathway <- NULL
names(MOX_filtered) <- c("Mean_rate", "Wetland_type", "Incubation_temperature", "Incubation_length", "Depth", "Salinity", "Dominant_vegetation", "Method", "Subpathway")
MOX_filtered$Subpathway <- NULL #no subpathways exist for MOx

wetland_type_levels_MOX <- levels(factor(MOX_filtered$Wetland_type))
dom_veg_levels_MOX <- levels(factor(MOX_filtered$Dominant_vegetation))
method_levels_MOX <- levels(factor(MOX_filtered$Method))

MOX_encoded <- MOX_filtered %>%
  mutate(
    Wetland_type = as.integer(factor(Wetland_type)),
    Dominant_vegetation = as.integer(factor(Dominant_vegetation)),
    Method = as.integer(factor(Method))
  )

#AOM
AOM_filtered=flux_rates_filtered[flux_rates_filtered$Pathway == "Anaerobic methane oxidation", ]
AOM_filtered$Pathway <- NULL
names(AOM_filtered) <- c("Mean_rate", "Wetland_type", "Incubation_temperature", "Incubation_length", "Depth", "Salinity", "Dominant_vegetation", "Method", "Subpathway")

wetland_type_levels_AOM <- levels(factor(AOM_filtered$Wetland_type))
dom_veg_levels_AOM <- levels(factor(AOM_filtered$Dominant_vegetation))
method_levels_AOM <- levels(factor(AOM_filtered$Method))
subpathway_levels_AOM <- levels(factor(AOM_filtered$Subpathway))

AOM_encoded <- AOM_filtered %>%
  mutate(
    Wetland_type = as.integer(factor(Wetland_type)),
    Dominant_vegetation = as.integer(factor(Dominant_vegetation)),
    Method = as.integer(factor(Method)),
    Subpathway = as.integer(factor(Subpathway))
  )

```

Data exploration 
```{r, warning=FALSE, message=FALSE}
library(GGally)

#Methanogenesis 

boxplot(MGEN_filtered$Mean_rate)
dotchart(MGEN_filtered$Mean_rate)

hist(MGEN_filtered$Mean_rate)
qqnorm(MGEN_filtered$Mean_rate)
qqline(MGEN_filtered$Mean_rate)

#numeric relationships 
ggpairs(MGEN_filtered[, sapply(MGEN_filtered, is.numeric)])
#categorical relationships 
boxplot(MGEN_filtered$Mean_rate ~ MGEN_filtered$Wetland_type)
boxplot(MGEN_filtered$Mean_rate ~ MGEN_filtered$Salinity)
boxplot(MGEN_filtered$Mean_rate ~ MGEN_filtered$Dominant_vegetation)
boxplot(MGEN_filtered$Mean_rate ~ MGEN_filtered$Method)
boxplot(MGEN_filtered$Mean_rate ~ MGEN_filtered$Subpathway)

#Aerobic methane oxidation

boxplot(MOX_filtered$Mean_rate)
dotchart(MOX_filtered$Mean_rate)

hist(MOX_filtered$Mean_rate)
qqnorm(MOX_filtered$Mean_rate)
qqline(MOX_filtered$Mean_rate)

#numeric relationships 
ggpairs(MOX_filtered[, sapply(MOX_filtered, is.numeric)])
#categorical relationships 
boxplot(MOX_filtered$Mean_rate ~ MOX_filtered$Wetland_type)
boxplot(MOX_filtered$Mean_rate ~ MOX_filtered$Salinity)
boxplot(MOX_filtered$Mean_rate ~ MOX_filtered$Dominant_vegetation)
boxplot(MOX_filtered$Mean_rate ~ MOX_filtered$Method)


#Anaerobic oxidation of methane
boxplot(AOM_filtered$Mean_rate)
dotchart(AOM_filtered$Mean_rate)

hist(AOM_filtered$Mean_rate)
qqnorm(AOM_filtered$Mean_rate)
qqline(AOM_filtered$Mean_rate)

#numeric relationships 
ggpairs(AOM_filtered[, sapply(AOM_filtered, is.numeric)])
#categorical relationships 
boxplot(AOM_filtered$Mean_rate ~ AOM_filtered$Wetland_type)
boxplot(AOM_filtered$Mean_rate ~ AOM_filtered$Salinity)
boxplot(AOM_filtered$Mean_rate ~ AOM_filtered$Dominant_vegetation)
boxplot(AOM_filtered$Mean_rate ~ AOM_filtered$Method)
boxplot(AOM_filtered$Mean_rate ~ AOM_filtered$Subpathway)

#Investigate confounding effect between wetland type and dominant vegetation with chi squared tests and by plotting proportions 

#methanogenesis
chisq_test_MGEN <- MGEN_filtered %>%
    filter(!is.na(Dominant_vegetation)) %>%
    with(table(Dominant_vegetation, Wetland_type)) %>%
    chisq.test()

print(chisq_test_MGEN)

library(forcats)

proportions_MGEN <- MGEN_filtered %>%
    filter(!is.na(Dominant_vegetation), !is.na(Wetland_type)) %>%
    count(Dominant_vegetation, Wetland_type) %>%
    group_by(Wetland_type) %>%
    mutate(Proportion = n / sum(n)) %>%
    ungroup()

ggplot(proportions_MGEN, aes(x = fct_reorder(Dominant_vegetation, Proportion), y = Proportion, fill = Wetland_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Methanogenesis",
         x = "Dominant Vegetation",
         y = "Proportion",
         fill = "Wetland Type")


#MOx
chisq_test_MOX <- MOX_filtered %>%
    filter(!is.na(Dominant_vegetation)) %>%
    with(table(Dominant_vegetation, Wetland_type)) %>%
    chisq.test()

print(chisq_test_MOX)

proportions_MOX <- MOX_filtered %>%
    filter(!is.na(Dominant_vegetation), !is.na(Wetland_type)) %>%
    count(Dominant_vegetation, Wetland_type) %>%
    group_by(Wetland_type) %>%
    mutate(Proportion = n / sum(n)) %>%
    ungroup()

ggplot(proportions_MOX, aes(x = fct_reorder(Dominant_vegetation, Proportion), y = Proportion, fill = Wetland_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Aerobic methane oxidation",
         x = "Dominant Vegetation",
         y = "Proportion",
         fill = "Wetland Type")

#AOM
chisq_test_AOM <- AOM_filtered %>%
    filter(!is.na(Dominant_vegetation)) %>%
    with(table(Dominant_vegetation, Wetland_type)) %>%
    chisq.test()

print(chisq_test_AOM)

proportions_AOM <- AOM_filtered %>%
    filter(!is.na(Dominant_vegetation), !is.na(Wetland_type)) %>%
    count(Dominant_vegetation, Wetland_type) %>%
    group_by(Wetland_type) %>%
    mutate(Proportion = n / sum(n)) %>%
    ungroup()

ggplot(proportions_AOM, aes(x = fct_reorder(Dominant_vegetation, Proportion), y = Proportion, fill = Wetland_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Anaerobic methane oxidation",
         x = "Dominant Vegetation",
         y = "Proportion",
         fill = "Wetland Type")



#count how many papers, unique sites, and unique measurements for each pathway 

#unique measurements:
flux_rates %>%
  filter(Pathway %in% c("Methanogenesis", "Aerobic methane oxidation", "Anaerobic methane oxidation")) %>%
  group_by(Pathway) %>%
  summarise(row_count = n())

# unique sites. Site is defined as unique Lat Long combination for a measurement. 
flux_rates %>%
  filter(Pathway %in% c("Methanogenesis", "Aerobic methane oxidation", "Anaerobic methane oxidation")) %>%
  group_by(Pathway) %>%
  summarise(unique_sites = n_distinct(paste0(Lat, "_", Long)))

#total unique sites.
flux_rates %>%
  distinct(Lat, Long) %>%
  summarise(unique_sites = n())

#unique studies.
flux_rates %>%
  filter(Pathway %in% c("Methanogenesis", "Aerobic methane oxidation", "Anaerobic methane oxidation")) %>%
  group_by(Pathway) %>%
  summarise(unique_studies = n_distinct(Study))

flux_rates %>%
summarise(unique_studies = n_distinct(Study))

```

Rate data significance testing and boxplot figure.
```{r}
library(dunn.test)
library(ggsignif)

#Kruskal-Wallis Test with post-hoc Dunn test.

# There was a statistically significant difference in rate between the different pathways (χ2(2)=150.12, p<2.2e-16). Significant differences were observed between MOx and AOM (Z = 11.06, adjusted p<0.00001), MGEN and AOM (Z = 9.82, adjusted p<0.00001), and MOx and MGEN (Z = 2.69, adjusted p=0.01).

kruskal.test(Mean_rate ~ Pathway, data = flux_rates_filtered)
dunn.test(flux_rates_filtered$Mean_rate, flux_rates_filtered$Pathway, method = "bonferroni")

#calculate median rate and IQR for each pathway
flux_rates_filtered %>%
  group_by(Pathway) %>%
  summarize(
    Median = median(Mean_rate, na.rm = TRUE),
    IQR = IQR(Mean_rate, na.rm = TRUE)
  )


## boxplot. Log transformed y axis. 
Fig2=ggplot(flux_rates, aes(x = Pathway, y = Mean_rate, fill = Pathway)) +
geom_boxplot(width = 0.7, outlier.shape = NA, alpha = 0.9) +  # Remove outlier highlighting
  scale_fill_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb")) +
  geom_jitter(color = "grey39", size = 1.5, alpha = 0.3) +
  geom_text(aes(label = "b", x = "Methanogenesis", y = 100), size = 4.5) +
  geom_text(aes(label = "a", x = "Aerobic methane oxidation", y = 100), size = 4.5) +
  geom_text(aes(label = "c", x = "Anaerobic methane oxidation", y = 100), size = 4.5) +
  labs(y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")"))) +
  scale_y_log10(
    breaks = c(1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1, 10, 100),  # Set custom breaks
    labels = c("0.00001", "0.0001", "0.001", "0.01", "0.1", "1", "10", "100")
  ) +
  annotation_logticks(sides = "l") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(color = "black"),  # Set color for x-axis text
    axis.text.y = element_text(color = "black"),  # Set color for y-axis text
    axis.title.y = element_text(color = "black"),  # Set color for y-axis title
    text = element_text(size = 12, color = "black")  # Set color for all text
  ) +
  xlab(NULL)


```

Map figure
```{r}
#function to convert coordinates from to degrees and decimal minutes to decimal degrees.

 convert_to_decimal <- function(coord) {
  # Split the coordinate into degrees, decimal minutes, and direction
  parts <- strsplit(coord, "_")[[1]]  # Split at "_"
  degrees <- as.numeric(parts[1])    # First part is degrees
  minutes <- as.numeric(substr(parts[2], 1, nchar(parts[2]) - 1))  # Extract decimal minutes
  direction <- substr(parts[2], nchar(parts[2]), nchar(parts[2]))  # Extract direction (N/S/E/W)
  
  # Convert to decimal degrees
  decimal <- degrees + (minutes / 60) 
  
  # Apply hemisphere: S/W are negative
  if (direction %in% c("S", "W")) {
    decimal <- -decimal
  }
  
  return(decimal)
}

flux_rates$Lat_Dec <- sapply(flux_rates$Lat, convert_to_decimal)
flux_rates$Long_Dec <- sapply(flux_rates$Long, convert_to_decimal)

# create a count column counting each coordinate grouped by process (this is only for plotting purposes)
library(dplyr)
flux_rates <- flux_rates %>%
  group_by(Lat_Dec, Long_Dec, Pathway) %>%
  mutate(count = n()) %>% 
  ungroup()


library(maps)

# Create a world map
world_map <- map_data("world")

## Map figure.
Fig1=ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "lightgray") +
  geom_path(data = world_map, aes(x = long, y = lat, group = group), 
            color = "black", size = 0.2, alpha = 0.1) + 
  geom_point(data = flux_rates, 
             aes(x = Long_Dec, y = Lat_Dec, 
                 size = count,               
                 fill = Pathway,            
                 shape = Pathway),          
             color = "black",             
             stroke = 0.6,                
             alpha = 0.9) +               
  scale_size_continuous(range = c(2, 8), name = "Measurements") + 
  scale_fill_manual(values = c("Methanogenesis" = "#66c2a5",
                                "Aerobic methane oxidation" = "#fc8d62",
                                "Anaerobic methane oxidation" = "#8da0cb")) + 
  scale_shape_manual(values = c("Methanogenesis" = 21,  
                                 "Aerobic methane oxidation" = 24, 
                                 "Anaerobic methane oxidation" = 22)) + 
  guides(size = "none",              
         shape = guide_legend(override.aes = list(alpha = 1, size = 4), nrow = 1, title = NULL),  
         fill = guide_legend(title = NULL, nrow = 1)) +  
  labs(x = "Longitude", y = "Latitude") +
  theme_minimal() +
  theme(text = element_text(size = 12, color = "black"),
        legend.position = "bottom", 
        legend.box = "horizontal",
        plot.margin = margin(t = 5, r = 5, b = 10, l = 5))  

```

Methanogenesis boosted regression tree model
```{r}
#This workflow combines the strengths of xgboost for model building and tidymodels for cross-validation, tuning, and organizing the workflow.

# Load necessary libraries
library(xgboost)
library(tidymodels)

# Set variables for model
data <- MGEN_encoded
response <- data$Mean_rate

predictors <- data %>% select(-Mean_rate)

# Grid search setup
set.seed(12)

# Model specification with early stopping
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune(),
  learn_rate = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Parameter grid
Xgb_grid <- grid_space_filling(
  trees(), tree_depth(), min_n(), loss_reduction(),
  sample_size = sample_prop(), finalize(mtry(), data),
  learn_rate(), size = 50
)

# Workflow setup
workflow <- workflow() %>%
  add_formula(Mean_rate ~ .) %>%
  add_model(xgb_spec)

# Cross-validation
set.seed(14) #was 12
folds <- vfold_cv(data, v = 5, repeats = 10) #was 2

# Model tuning
doParallel::registerDoParallel()
set.seed(14) #was 12
Xgb_res <- tune_grid(
  workflow, resample = folds, grid = Xgb_grid,
  control = control_grid(save_pred = TRUE)
)

# Extract best model parameters
Best_rmse <- select_best(Xgb_res)

#This object contains the final model parameters
final_xgb <- finalize_workflow(workflow, Best_rmse)

# Final model and variable importance. This object contains even more hyperparameters (under params)
final_xgb_fit <- final_xgb %>%
  fit(data = data)

# Extract the trained model and plot variable importance
xgb_model <- extract_fit_engine(final_xgb_fit)
importance_matrix <- xgb.importance(model = xgb_model)

# Convert the importance matrix to a dataframe
importance_df <- as.data.frame(importance_matrix)

# Convert Gain scores to percentages
importance_df$Gain_percentage <- importance_df$Gain * 100

# Plotting the variable importance with Gain percentage as the x-axis
ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance", x = "Features", y = "Importance Score (%)")


final_fit_with_metrics <- final_xgb %>%
  fit_resamples(
    resamples = folds,
    control = control_resamples(save_pred = TRUE, save_workflow = TRUE)
  )

# Collect metrics from the model. RMSE/R2 for model validation table.
metrics <- final_fit_with_metrics %>%
  collect_metrics()

#Let's compare model metrics to the range and sd of my response variable
range(data$Mean_rate)
sd(data$Mean_rate)

#CV learning curves. Plot to tune number of trees, alpha and lambda values. 

# Step 1: Perform cross-validation with the full data
set.seed(123)
folds <- vfold_cv(data, v = 5, repeats = 10)


num_columns <- ncol(data) - 1 
params <- list(
  colsample_bytree = min(1, 5 / num_columns),
  colsample_bynode = 0.125, 
  min_child_weight = 6, 
  max_depth = 10, 
  eta = 0.0184206996932672, 
  gamma = 0.00954095476349994, 
  subsample = 0.46734693877551, 
  lambda = 2, 
  alpha = 3 
   
)

# Step 2: Create learning curve data structure
train_sizes <- seq(0.4, 1.0, by = 0.1)
learning_curve_data <- data.frame()

# Loop over each fold in cross-validation
for (fold in folds$splits) {
 
  # Extract training and testing data for this fold
  train_data <- analysis(fold)
  test_data <- assessment(fold)
 
  # Prepare the test data for xgboost (held constant for evaluation)
  test_x <- data.matrix(test_data[, -which(colnames(test_data) == "Mean_rate")])
  test_y <- test_data[["Mean_rate"]]
  dtest <- xgb.DMatrix(data = test_x, label = test_y)
 
  # Step 3: Iterate over training sizes
  for (size in train_sizes) {
   
    # Create a subset of the training data
    train_subset <- sample_frac(train_data, size = size)
   
    # Prepare the subset for xgboost
    train_x <- data.matrix(train_subset[, -which(colnames(train_subset) == "Mean_rate")])
    train_y <- train_subset[["Mean_rate"]]
    dtrain <- xgb.DMatrix(data = train_x, label = train_y)
   
    # Train the model using the subset
    xgb_model <- xgb.train(
      params = params,
      data = dtrain,
      nrounds = 10000,
      watchlist = list(train = dtrain, test = dtest),
      early_stopping_rounds = NULL,
      eval_metric = "rmse",
      verbose = 0,
    )
   
    # Extract evaluation log for this subset
    evals <- xgb_model$evaluation_log
    final_train_rmse <- tail(evals$train_rmse, 1)
    final_test_rmse <- tail(evals$test_rmse, 1)
   
    # Store results in learning curve data
    learning_curve_data <- rbind(
      learning_curve_data,
      data.frame(fold = as.numeric(fold$id),
                 size = size,
                 train_rmse = final_train_rmse,
                 test_rmse = final_test_rmse)
    )
  }
}



evaluation_log <- xgb_model$evaluation_log

# Plot the learning curve

MGEN_LC=ggplot(evaluation_log, aes(x = iter)) +
  geom_line(aes(y = train_rmse, color = 'Train')) +
  geom_line(aes(y = test_rmse, color = 'Test')) +
  geom_vline(xintercept = 9000, linetype = "dashed", color = "black") +  
  labs(y = 'RMSE', x = 'Iterations', color = 'Dataset',
       title = 'Methanogenesis', 
       tag = "a)") +
  scale_color_manual(values = c('Train' = '#1f78b4', 'Test' = '#f26928')) + 
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))


# Fit final model with all tuned hyperparameters

# Define cross-validation folds
set.seed(14) #was 12
folds <- vfold_cv(data, v = 5, repeats = 10)

# Specify final model parameters
num_columns <- ncol(data) - 1  
final_params <- list(
  colsample_bytree = min(1, 5 / num_columns),
  colsample_bynode = 0.125,
  min_child_weight = 6,
  max_depth = 10,
  eta = 0.0184206996932672, 
  gamma = 0.00954095476349994, 
  subsample = 0.46734693877551, 
  lambda = 2, 
  alpha = 3 
   
)

# Final XGBoost model specification
xgb_spec_final <- boost_tree(
  trees = 9000,                        
  tree_depth = 10,                      
  min_n = 6,                           
  loss_reduction = 0.00954095476349994, 
  sample_size = 0.46734693877551,     
  mtry = 1,                            
  learn_rate = 0.0184206996932672      
) %>%
  set_engine("xgboost", lambda = 2, alpha = 3) %>% 
  set_mode("regression")

# Workflow setup. Hyperparameters.
workflow_final <- workflow() %>%
  add_formula(Mean_rate ~ .) %>%
  add_model(xgb_spec_final)

# Model fitting and resampling
final_fit_with_metrics <- workflow_final %>%
  fit_resamples(
    resamples = folds,
    control = control_resamples(save_pred = TRUE, save_workflow = TRUE)
  )

# Collect model metrics (e.g., RMSE, R-squared)
metrics_final <- final_fit_with_metrics %>%
  collect_metrics()

#Calculate AIC
# Extract predictions from the final model
preds <- predict(final_xgb_fit, new_data = data)$.pred

# Compute residual sum of squares (RSS)
rss <- sum((data$Mean_rate - preds)^2)

# Number of observations
n <- nrow(data)

# Number of parameters (approximate as number of trees)
k <- 9000 

# Compute AIC
aic_xgb <- n * log(rss / n) + 2 * k



# AIC=nlog(RMSE)+2p (Shabani 2017)

# Get number of observations
n <- nrow(MGEN_encoded)

# Extract RMSE from collected metrics
RMSE_value <- 1.27

# Define number of explanatory variables (excluding response variable)
P <- ncol(MGEN_encoded) - 1  

# Compute AIC
AIC_value <- n * log(RMSE_value) + 2 * P

# Print result
AIC_value



# Fit the final model on the full dataset. 
final_xgb_fit <- workflow_final %>%
  fit(data = data)

# Extract the trained model and plot variable importance
xgb_model_final <- extract_fit_engine(final_xgb_fit)
importance_matrix_final <- xgb.importance(model = xgb_model_final)

# Convert the importance matrix to a dataframe and plot
importance_df <- as.data.frame(importance_matrix_final)
importance_df$Gain_percentage <- importance_df$Gain * 100

# Plot variable importance
ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance", x = "Features", y = "Importance Score (%)") +
  theme_minimal()


## Feature importance plots.
triMGEN=ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage, fill = Feature)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Methanogenesis",
       x = "Features",
       y = "Relative Importance (%)",
       tag = "a)") +
  scale_x_discrete(labels = c("Salinity","Wetland type","Method","Pathway","Incubation temperature","Incubation length","Dominant vegetation","Depth")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Pathway" = "#B3B3B3", 
                               "Method"="#B3B3B3",
                               "Wetland_type"="#B3B3B3",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#B3B3B3")) +
  theme(axis.text.y = element_text(hjust = 1, color = c("#999999","#999999","#999999","#999999","#e41a1c","#999999", "#4daf4a", "#a65628")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none", text = element_text(size = 12))


##partial dependency plots. 

library(pdp)

predictor_names <- colnames(predictors)


# Loop through each predictor to generate PDPs
for (pred in predictor_names) {
  pdp_result <- partial(
    xgb_model_final, 
    pred.var = pred, 
    train = as.matrix(predictors),
    plot = FALSE
  )
  
  # Convert the partial dependence result to a dataframe
  pdp_df <- as.data.frame(pdp_result)
  
  # Plot using ggplot2
  p <- ggplot(pdp_df, aes_string(x = pred, y = "yhat")) +
    geom_line(color = "steelblue", size = 1) +
    labs(
      title = paste("Partial Dependency of", pred),
      x = pred,
      y = "Predicted Mean Rate"
    ) +
    theme_minimal()
  
  # Print each plot
  print(p)
}


# Partial dependence plot for Depth
pdp_depth <- partial(
  xgb_model_final, 
  pred.var = "Depth", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_depth_df <- as.data.frame(pdp_depth)

# Plot using ggplot2 for Depth
MGEN_pdp_depth <- ggplot(pdp_depth_df, aes_string(x = "Depth", y = "yhat")) +
      geom_line() +
    geom_rug(data = MGEN_filtered, aes(x = Depth, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Depth (cm)", y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")), tag = "g)") +
    theme_minimal()


# Partial dependence plot for Incubation Temperature
pdp_temp <- partial(
  xgb_model_final, 
  pred.var = "Incubation_temperature", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_temp_df <- as.data.frame(pdp_temp)

# Plot using ggplot2 for Incubation Temperature
MGEN_pdp_temp <- ggplot(pdp_temp_df, aes_string(x = "Incubation_temperature", y = "yhat")) +
    geom_line() +
    geom_rug(data = MGEN_filtered, aes(x = Incubation_temperature, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Incubation Temperature (\u00b0C)", y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")),tag = "d)") +
    theme_minimal()


# Partial dependence plot for Dominant Vegetation
pdp_domveg <- partial(
  xgb_model_final, 
  pred.var = "Dominant_vegetation", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_domveg_df <- as.data.frame(pdp_domveg)

# map the species integers back to their categories for plotting
species_list <- c(
  "Bare mudflat", "Caladium jamaicense", "Carex sp.", "Cyperus malacensis", 
  "Cyrilla racemiflora", "Equisetum variegatum", "Fraxinus caroliniana", 
  "Hippuris vulgaris", "Juncus effuses", "Lythrum salicaria", 
  "Menyanthes trifoliata", "Nuphar polysepalum", "Phragmites australis", 
  "Spartina alterniflora", "Spartina mariqueter", "Sphagnum sp.", "Typha sp."
)

# Remove the first row from the dataframe
pdp_domveg_df <- pdp_domveg_df[-1, ]

# Assuming 'Dominant_vegetation' column exists in pdp_domveg_df and is ranked 1-17
pdp_domveg_df <- pdp_domveg_df %>%
  mutate(species = species_list[Dominant_vegetation])


# Plot using ggplot2 for Dom veg
MGEN_pdp_domveg <- ggplot(pdp_domveg_df, aes(x = factor(species, levels = species[order(yhat)]), y = yhat)) +
  geom_point(size = 8, shape = 95) +
  labs(
    title = "Methanogenesis",
    x = "Dominant Vegetation", 
    y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")),
    tag = "a)") + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )



```

Aerobic methane oxidation boosted regression tree model
```{r}
#This workflow combines the strengths of xgboost for model building and tidymodels for cross-validation, tuning, and organizing the workflow.

# Load necessary libraries
library(xgboost)
library(tidymodels)

# Set variables for model
data <- MOX_encoded
response <- data$Mean_rate
predictors <- data %>% select(-Mean_rate)

# Grid search setup
set.seed(362)

# Model specification with early stopping
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune(),
  learn_rate = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Parameter grid
Xgb_grid <- grid_space_filling(
  trees(), tree_depth(), min_n(), loss_reduction(),
  sample_size = sample_prop(), finalize(mtry(), data),
  learn_rate(), size = 50
)

# Workflow setup
workflow <- workflow() %>%
  add_formula(Mean_rate ~ .) %>%
  add_model(xgb_spec)

# Cross-val1dation
set.seed(376)
folds <- vfold_cv(data, v = 5, repeats = 10)

# Model tuning
doParallel::registerDoParallel()
set.seed(376)

Xgb_res <- tune_grid(
  workflow, resample = folds, grid = Xgb_grid,
  control = control_grid(save_pred = TRUE)
)

# Extract best model parameters
Best_rmse <- select_best(Xgb_res)

#This object contains the final model parameters
final_xgb <- finalize_workflow(workflow, Best_rmse)

# Final model and variable importance. 
final_xgb_fit <- final_xgb %>%
  fit(data = data)

# Extract the trained model and plot variable importance
xgb_model <- extract_fit_engine(final_xgb_fit)
importance_matrix <- xgb.importance(model = xgb_model)

# Convert the importance matrix to a dataframe
importance_df <- as.data.frame(importance_matrix)

# Convert Gain scores to percentages
importance_df$Gain_percentage <- importance_df$Gain * 100

# Plotting the variable importance with Gain percentage as the x-axis
ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance", x = "Features", y = "Importance Score (%)")


final_fit_with_metrics <- final_xgb %>%
  fit_resamples(
    resamples = folds,
    control = control_resamples(save_pred = TRUE, save_workflow = TRUE)
  )

# Collect metrics from the model. RMSE/R2 for model validation table.
metrics <- final_fit_with_metrics %>%
  collect_metrics()

#Let's compare model metrics to the range and sd of my response variable
range(data$Mean_rate)
sd(data$Mean_rate)

#CV learning curves. Plot to tune number of trees, alpha and lambda values. 

# Step 1: Perform cross-validation with the full data
set.seed(456)
folds <- vfold_cv(data, v = 5, repeats = 10)

num_columns <- ncol(data) - 1 
params <- list(
  colsample_bytree = min(1, 5 / num_columns),
  colsample_bynode = 0.142857142857143,
  min_child_weight = 6,
  max_depth = 10, #was 10
  eta = 0.0184206996932672,
  gamma = 0.00954095476349994,
  subsample = 0.46734693877551,
  lambda = 4, 
  alpha = 8 
   
)

# Step 2: Create learning curve data structure
train_sizes <- seq(0.4, 1.0, by = 0.1)  
learning_curve_data <- data.frame()

# Loop over each fold in cross-validation
for (fold in folds$splits) {
 
  # Extract training and testing data for this fold
  train_data <- analysis(fold)
  test_data <- assessment(fold)
 
  # Prepare the test data for xgboost (held constant for evaluation)
  test_x <- data.matrix(test_data[, -which(colnames(test_data) == "Mean_rate")])
  test_y <- test_data[["Mean_rate"]]
  dtest <- xgb.DMatrix(data = test_x, label = test_y)
 
  # Step 3: Iterate over training sizes
  for (size in train_sizes) {
   
    # Create a subset of the training data
    train_subset <- sample_frac(train_data, size = size)
   
    # Prepare the subset for xgboost
    train_x <- data.matrix(train_subset[, -which(colnames(train_subset) == "Mean_rate")])
    train_y <- train_subset[["Mean_rate"]]
    dtrain <- xgb.DMatrix(data = train_x, label = train_y)
   
    # Train the model using the subset
    xgb_model <- xgb.train(
      params = params,
      data = dtrain,
      nrounds = 10000,
      watchlist = list(train = dtrain, test = dtest),
      early_stopping_rounds = NULL,
      eval_metric = "rmse",
      verbose = 0,
    )
   
    # Extract evaluation log for this subset
    evals <- xgb_model$evaluation_log
    final_train_rmse <- tail(evals$train_rmse, 1)
    final_test_rmse <- tail(evals$test_rmse, 1)
   
    # Store results in learning curve data
    learning_curve_data <- rbind(
      learning_curve_data,
      data.frame(fold = as.numeric(fold$id),
                 size = size,
                 train_rmse = final_train_rmse,
                 test_rmse = final_test_rmse)
    )
  }
}



evaluation_log <- xgb_model$evaluation_log

# Plot the learning curve

MOX_LC=ggplot(evaluation_log, aes(x = iter)) +
  geom_line(aes(y = train_rmse, color = 'Train')) +
  geom_line(aes(y = test_rmse, color = 'Test')) +
  geom_vline(xintercept = 8500, linetype = "dashed", color = "black") +  
  labs(y = 'RMSE', x = 'Iterations', color = 'Dataset',
       title = 'Aerobic methane oxidation',
       tag = "b)") +
  scale_color_manual(values = c('Train' = '#1f78b4', 'Test' = '#f26928')) + 
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))


# Fit final model with all tuned hyperparameters.

# Define cross-validation folds
set.seed(4013) #was 4002
folds <- vfold_cv(data, v = 5, repeats = 10)

  
# Specify final model parameters
num_columns <- ncol(data) - 1  
final_params <- list(
  colsample_bytree = min(1, 5 / num_columns),
  colsample_bynode = 0.142857142857143,
  min_child_weight = 6,
  max_depth = 10,
  eta = 0.0184206996932672, 
  gamma = 0.00954095476349994,
  subsample = 0.46734693877551                 
)

# Final XGBoost model specification
xgb_spec_final <- boost_tree(
  trees = 8500,                        
  tree_depth = 10,                      
  min_n = 6,                           
  loss_reduction = 0.00954095476349994, 
  sample_size = 0.46734693877551,     
  mtry = 1,                           
  learn_rate = 0.0184206996932672      
) %>%
    set_engine("xgboost", lambda = 4, alpha = 8) %>% 
  set_mode("regression")


# Workflow setup. Hyperparameters.
workflow_final <- workflow() %>%
  add_formula(Mean_rate ~ .) %>%
  add_model(xgb_spec_final)

# Model fitting and resampling
final_fit_with_metrics <- workflow_final %>%
  fit_resamples(
    resamples = folds,
    control = control_resamples(save_pred = TRUE, save_workflow = TRUE)
  )

# Collect model metrics (e.g., RMSE, R-squared)
metrics_final <- final_fit_with_metrics %>%
  collect_metrics()


# Fit the final model on the full dataset. 
final_xgb_fit <- workflow_final %>%
  fit(data = data)

#Calculate AIC
# Extract predictions from the final model
preds <- predict(final_xgb_fit, new_data = data)$.pred

# Compute residual sum of squares (RSS)
rss <- sum((data$Mean_rate - preds)^2)

# Number of observations
n <- nrow(data)

# Number of parameters (approximate as number of trees)
k <- 8500 

# Compute AIC
aic_xgb <- n * log(rss / n) + 2 * k

# AIC=nlog(RMSE)+2p (Shabani 2017)

# Get number of observations
n <- nrow(MOX_encoded)

# Extract RMSE from collected metrics
RMSE_value <- 2.76

# Define number of explanatory variables (excluding response variable)
P <- ncol(MOX_encoded) - 1  

# Compute AIC
AIC_value <- n * log(RMSE_value) + 2 * P

# Print result
AIC_value

# Extract the trained model and plot variable importance
xgb_model_final <- extract_fit_engine(final_xgb_fit)
importance_matrix_final <- xgb.importance(model = xgb_model_final)

# Convert the importance matrix to a dataframe and plot
importance_df <- as.data.frame(importance_matrix_final)
importance_df$Gain_percentage <- importance_df$Gain * 100

# Plot variable importance
ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance", x = "Features", y = "Importance Score (%)") +
  theme_minimal()


## feature importance plot.
triMOX=ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage, fill = Feature)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Aerobic methane oxidation",
       x = "Features",
       y = "Relative Importance (%)",
       tag = "b)") +
  scale_x_discrete(labels = c("Salinity","Method","Wetland type","Incubation temperature","Incubation length","Depth","Dominant vegetation")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Method"="#B3B3B3",
                               "Wetland_type"="#B3B3B3",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#B3B3B3")) +
  theme(axis.text.y = element_text(hjust = 1, color = c("#999999","#999999","#999999","#e41a1c","#999999","#a65628","#4daf4a")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none", text = element_text(size = 12))


##partial dependency plots. 

library(pdp)

predictor_names <- colnames(predictors)


# Loop through each predictor to generate PDPs
for (pred in predictor_names) {
  pdp_result <- partial(
    xgb_model_final, 
    pred.var = pred, 
    train = as.matrix(predictors),
    plot = FALSE
  )
  
  # Convert the partial dependence result to a dataframe
  pdp_df <- as.data.frame(pdp_result)
  
  # Plot using ggplot2
  p <- ggplot(pdp_df, aes_string(x = pred, y = "yhat")) +
    geom_line(color = "steelblue", size = 1) +
    labs(
      title = paste("Partial Dependency of", pred),
      x = pred,
      y = "Predicted Mean Rate"
    ) +
    theme_minimal()
  
  # Print each plot
  print(p)
}


# Partial dependence plot for Depth
pdp_depth <- partial(
  xgb_model_final, 
  pred.var = "Depth", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_depth_df <- as.data.frame(pdp_depth)

# Plot using ggplot2 for Depth
MOX_pdp_depth <- ggplot(pdp_depth_df, aes_string(x = "Depth", y = "yhat")) +
      geom_line() +
    geom_rug(data = MOX_filtered, aes(x = Depth, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Depth (cm)",  y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")),tag = "h)") +
    theme_minimal()

# Print the Depth plot
print(MOX_pdp_depth)


# Partial dependence plot for Incubation Temperature
pdp_temp <- partial(
  xgb_model_final, 
  pred.var = "Incubation_temperature", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_temp_df <- as.data.frame(pdp_temp)

# Plot using ggplot2 for Incubation Temperature
MOX_pdp_temp <- ggplot(pdp_temp_df, aes_string(x = "Incubation_temperature", y = "yhat")) +
    geom_line() +
    geom_rug(data = MOX_filtered, aes(x = Incubation_temperature, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Incubation Temperature (\u00b0C)",  y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")),tag = "e)") +
    theme_minimal()

# Print the Incubation Temperature plot
print(MOX_pdp_temp)

# Partial dependence plot for Dominant Vegetation
pdp_domveg <- partial(
  xgb_model_final, 
  pred.var = "Dominant_vegetation", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_domveg_df <- as.data.frame(pdp_domveg)

# map the species integers back to their categories for plotting
species_list <- c(
  "Bare mudflat", "Carex sp.", "Cyrilla racemiflora", "Eriophorum angustifolium", "Fraxinus caroliniana", "Oryza sp.","Phragmites australis", "Spartina alterniflora", "Spartina mariqueter", "Sphagnum sp.", "Typha sp."
)

# Remove the first row from the dataframe
pdp_domveg_df <- pdp_domveg_df[-1, ]

# Assuming 'Dominant_vegetation' column exists in pdp_domveg_df and is ranked 1-9
pdp_domveg_df <- pdp_domveg_df %>%
  mutate(species = species_list[Dominant_vegetation])


# Plot using ggplot2 for Dom veg
MOX_pdp_domveg <- ggplot(pdp_domveg_df, aes_string(x = "reorder(species, yhat)", y = "yhat")) +
    geom_point(size = 8, shape = 95) +
    labs(title = "Aerobic methane oxidation",
         x = "Dominant Vegetation", 
          y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")), tag = "b)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))


# Print the Dom veg plot
print(MOX_pdp_domveg)


```

Anaerobic methane oxidation boosted regression tree model
```{r}
#This workflow combines the strengths of xgboost for model building and tidymodels for cross-validation, tuning, and organizing the workflow.

# Load necessary libraries
library(xgboost)
library(tidymodels)

# Set variables for model
data <- AOM_encoded
response <- data$Mean_rate
predictors <- data %>% select(-Mean_rate)

# Grid search setup
set.seed(57)

# Model specification with early stopping
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune(),
  learn_rate = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Parameter grid
Xgb_grid <- grid_space_filling(
  trees(), tree_depth(), min_n(), loss_reduction(),
  sample_size = sample_prop(), finalize(mtry(), data),
  learn_rate(), size = 50
)

# Workflow setup
workflow <- workflow() %>%
  add_formula(Mean_rate ~ .) %>%
  add_model(xgb_spec)

# Cross-validation
set.seed(914)
folds <- vfold_cv(data, v = 5, repeats = 10) #was 5


# Model tuning
doParallel::registerDoParallel()
set.seed(914)
Xgb_res <- tune_grid(
  workflow, resample = folds, grid = Xgb_grid,
  control = control_grid(save_pred = TRUE)
)

# Extract best model parameters
Best_rmse <- select_best(Xgb_res)

#This object contains the final model parameters
final_xgb <- finalize_workflow(workflow, Best_rmse)

# Final model and variable importance. 
final_xgb_fit <- final_xgb %>%
  fit(data = data)

# Extract the trained model and plot variable importance
xgb_model <- extract_fit_engine(final_xgb_fit)
importance_matrix <- xgb.importance(model = xgb_model)

# Convert the importance matrix to a dataframe
importance_df <- as.data.frame(importance_matrix)

# Convert Gain scores to percentages
importance_df$Gain_percentage <- importance_df$Gain * 100

# Plotting the variable importance with Gain percentage as the x-axis
ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance", x = "Features", y = "Importance Score (%)")


final_fit_with_metrics <- final_xgb %>%
  fit_resamples(
    resamples = folds,
    control = control_resamples(save_pred = TRUE, save_workflow = TRUE)
  )

# Collect metrics from the model. RMSE/R2 for model validation table.
metrics <- final_fit_with_metrics %>%
  collect_metrics()

#Let's compare model metrics to the range and sd of my response variable
range(data$Mean_rate)
sd(data$Mean_rate)

#CV learning curves. Plot to tune number of trees, alpha and lambda values. 

# Step 1: Perform cross-validation with the full data
set.seed(301)
folds <- vfold_cv(data, v = 5, repeats = 10)

num_columns <- ncol(data) - 1 
params <- list(
  colsample_bytree = min(1, 5 / num_columns),
  colsample_bynode = 0.125,
  min_child_weight = 6,
  max_depth = 10,
  eta = 0.0184206996932672,
  gamma = 0.00954095476349994,
  subsample = 0.46734693877551
  #lambda = 4, #adjusting lambda did not improve the learning curves 
  #alpha = 8 #adjusting alpha did not improve the learning curves 
   
)

# Step 2: Create learning curve data structure
train_sizes <- seq(0.4, 1.0, by = 0.1)  
learning_curve_data <- data.frame()

# Loop over each fold in cross-validation
for (fold in folds$splits) {
 
  # Extract training and testing data for this fold
  train_data <- analysis(fold)
  test_data <- assessment(fold)
 
  # Prepare the test data for xgboost (held constant for evaluation)
  test_x <- data.matrix(test_data[, -which(colnames(test_data) == "Mean_rate")])
  test_y <- test_data[["Mean_rate"]]
  dtest <- xgb.DMatrix(data = test_x, label = test_y)
 
  # Step 3: Iterate over training sizes
  for (size in train_sizes) {
   
    # Create a subset of the training data
    train_subset <- sample_frac(train_data, size = size)
   
    # Prepare the subset for xgboost
    train_x <- data.matrix(train_subset[, -which(colnames(train_subset) == "Mean_rate")])
    train_y <- train_subset[["Mean_rate"]]
    dtrain <- xgb.DMatrix(data = train_x, label = train_y)
   
    # Train the model using the subset
    xgb_model <- xgb.train(
      params = params,
      data = dtrain,
      nrounds = 10000,
      watchlist = list(train = dtrain, test = dtest),
      early_stopping_rounds = NULL,
      eval_metric = "rmse",
      verbose = 0,
    )
   
    # Extract evaluation log for this subset
    evals <- xgb_model$evaluation_log
    final_train_rmse <- tail(evals$train_rmse, 1)
    final_test_rmse <- tail(evals$test_rmse, 1)
   
    # Store results in learning curve data
    learning_curve_data <- rbind(
      learning_curve_data,
      data.frame(fold = as.numeric(fold$id),
                 size = size,
                 train_rmse = final_train_rmse,
                 test_rmse = final_test_rmse)
    )
  }
}



evaluation_log <- xgb_model$evaluation_log

# Plot the learning curve

AOM_LC=ggplot(evaluation_log, aes(x = iter)) +
  geom_line(aes(y = train_rmse, color = 'Train')) +
  geom_line(aes(y = test_rmse, color = 'Test')) +
  geom_vline(xintercept = 7500, linetype = "dashed", color = "black") +  
  labs(y = 'RMSE', x = 'Iterations', color = 'Dataset',
       title = 'Anaerobic methane oxidation', 
       tag = "c)") +
  scale_color_manual(values = c('Train' = '#1f78b4', 'Test' = '#f26928')) +  
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))


# Fit final model with all tuned hyperparameters

# Define cross-validation folds
set.seed(418)
folds <- vfold_cv(data, v = 5, repeats = 10)

# Specify final model parameters
num_columns <- ncol(data) - 1  # Count features only
final_params <- list(
  colsample_bytree = min(1, 5 / num_columns),
  colsample_bynode = 0.125,
  min_child_weight = 6,
  max_depth = 10,
  eta = 0.0184206996932672,
  gamma = 0.00954095476349994,
  subsample = 0.46734693877551
)

# Final XGBoost model specification
xgb_spec_final <- boost_tree(
  trees = 7500,                        
  tree_depth = 10,                      
  min_n = 6,                           
  loss_reduction = 0.00954095476349994, 
  sample_size = 0.46734693877551,     
  mtry = 1,                            
  learn_rate = 0.0184206996932672     
) %>%
  set_engine("xgboost") %>%  
  set_mode("regression")

# Workflow setup. Hyperparameters.
workflow_final <- workflow() %>%
  add_formula(Mean_rate ~ .) %>%
  add_model(xgb_spec_final)

# Model fitting and resampling
final_fit_with_metrics <- workflow_final %>%
  fit_resamples(
    resamples = folds,
    control = control_resamples(save_pred = TRUE, save_workflow = TRUE)
  )

# Collect model metrics (e.g., RMSE, R-squared)
metrics_final <- final_fit_with_metrics %>%
  collect_metrics()


# Fit the final model on the full dataset. Here are the hyperparameters.
final_xgb_fit <- workflow_final %>%
  fit(data = data)

#Calculate AIC

#RSS method. AIC=nlog(rss/n) + 2k

# Extract predictions from the final model
preds <- predict(final_xgb_fit, new_data = data)$.pred

# Compute residual sum of squares (RSS)
rss <- sum((data$Mean_rate - preds)^2)

# Number of observations
n <- nrow(data)

# Number of parameters (approximate as number of trees)
k <- 7500

# Compute AIC
aic_xgb <- n * log(rss / n) + 2 * k

# RMSE method. AIC=nlog(RMSE)+2p (Shabani 2017)

# Get number of observations
n <- nrow(AOM_encoded)

# Extract RMSE from collected metrics
RMSE_value <- 0.21

# Define number of explanatory variables (excluding response variable)
P <- ncol(AOM_encoded) - 1  

# Compute AIC
AIC_value <- n * log(RMSE_value) + 2 * P

# Print result
AIC_value

# Extract the trained model and plot variable importance
xgb_model_final <- extract_fit_engine(final_xgb_fit)
importance_matrix_final <- xgb.importance(model = xgb_model_final)

# Convert the importance matrix to a dataframe and plot
importance_df <- as.data.frame(importance_matrix_final)

importance_df$Gain_percentage <- importance_df$Gain * 100

# Plot variable importance
ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable Importance", x = "Features", y = "Importance Score (%)") +
  theme_minimal()

## feature importance plot.
triAOM=ggplot(importance_df, aes(x = reorder(Feature, Gain_percentage), y = Gain_percentage, fill = Feature)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Anaerobic methane oxidation",
       x = "Features",
       y = "Relative Importance (%)",
       tag = "c)") +
  scale_x_discrete(labels = c("Method","Wetland type","Salinity","Incubation length","Pathway","Depth","Dominant vegetation","Incubation temperature")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Pathway" = "#B3B3B3", 
                               "Method"="#B3B3B3",
                               "Wetland_type"="#B3B3B3",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#B3B3B3")) +
  theme(axis.text.y = element_text(hjust = 1, color = c("#999999","#999999","#999999","#999999","#999999","#a65628","#4daf4a","#e41a1c")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none", text = element_text(size = 12))

##partial dependency plots. 

library(pdp)

predictor_names <- colnames(predictors)


# Loop through each predictor to generate PDPs
for (pred in predictor_names) {
  pdp_result <- partial(
    xgb_model_final, 
    pred.var = pred, 
    train = as.matrix(predictors),
    plot = FALSE
  )
  
  # Convert the partial dependence result to a dataframe
  pdp_df <- as.data.frame(pdp_result)
  
  # Plot using ggplot2
  p <- ggplot(pdp_df, aes_string(x = pred, y = "yhat")) +
    geom_line(color = "steelblue", size = 1) +
    labs(
      title = paste("Partial Dependency of", pred),
      x = pred,
      y = "Predicted Mean Rate"
    ) +
    theme_minimal()
  
  # Print each plot
  print(p)
}


# Partial dependence plot for Depth
pdp_depth <- partial(
  xgb_model_final, 
  pred.var = "Depth", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_depth_df <- as.data.frame(pdp_depth)

# Plot using ggplot2 for Depth
AOM_pdp_depth <- ggplot(pdp_depth_df, aes_string(x = "Depth", y = "yhat")) +
      geom_line() +
    geom_rug(data = AOM_filtered, aes(x = Depth, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Depth (cm)",  y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")), tag = "i)") +
    theme_minimal()

# Print the Depth plot
print(AOM_pdp_depth)


# Partial dependence plot for Incubation Temperature
pdp_temp <- partial(
  xgb_model_final, 
  pred.var = "Incubation_temperature", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_temp_df <- as.data.frame(pdp_temp)

# Plot using ggplot2 for Incubation Temperature
AOM_pdp_temp <- ggplot(pdp_temp_df, aes_string(x = "Incubation_temperature", y = "yhat")) +
    geom_line() +
    geom_rug(data = AOM_filtered, aes(x = Incubation_temperature, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Incubation Temperature (\u00b0C)",  y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")), tag = "f)") +
    theme_minimal()

# Print the Incubation Temperature plot
print(AOM_pdp_temp)

# Partial dependence plot for Dominant Vegetation
pdp_domveg <- partial(
  xgb_model_final, 
  pred.var = "Dominant_vegetation", 
  train = as.matrix(predictors),
  plot = FALSE
)

# Convert the partial dependence result to a dataframe
pdp_domveg_df <- as.data.frame(pdp_domveg)

# map the species integers back to their categories for plotting

species_list <- c("Bare mudflat", "Cyrilla racemiflora","Oryza sp.","Phragmites australis","Spartina alterniflora", "Spartina mariqueter", "Sphagnum sp.", "Suaeda glauca", "Tamarix sp.", "Unidentified reed")

species_df <- data.frame(species = species_list)

# Remove the first row from the dataframe
pdp_domveg_df <- pdp_domveg_df[-nrow(pdp_domveg_df), ]

# Assuming 'Dominant_vegetation' column exists in pdp_domveg_df and is ranked 1-9
pdp_domveg_df <- cbind(pdp_domveg_df, species = species_df$species)


# Plot using ggplot2 for Dom veg
AOM_pdp_domveg <- ggplot(pdp_domveg_df, aes_string(x = "reorder(species, yhat)", y = "yhat")) +
    geom_point(size = 8, shape = 95) +
    labs(title = "Anaerobic methane oxidation",
         x = "Dominant vegetation",  y = expression(paste("Rate (", mu, "mol CH"[4], " g"^{-1}, "soil d"^{-1}, ")")), tag = "c)") +
    theme_minimal()+
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))


# Print the Dom veg plot
print(AOM_pdp_domveg)


```

Final figures for paper
```{r}
library(ggpubr)

#Main paper. Saved as .tiff files with dpi=1000

# Fig 1
ggsave("Fig1.tiff", plot = Fig1, device = "tiff", dpi = 800, width = 8, height = 5.3, bg="white")

# Fig 2
ggsave("Fig2.tiff", plot = Fig2, device = "tiff", dpi = 800, width = 7, height = 6, bg="white")

# Fig 3
Fig3=ggarrange(triMGEN, triMOX, triAOM, ncol = 1, nrow = 3)

ggsave("Fig3.tiff", plot = Fig3, device = "tiff", dpi = 800, width = 6, height = 10, bg="white")

# Fig 4
Fig4=ggarrange(MGEN_pdp_domveg,MOX_pdp_domveg, AOM_pdp_domveg, MGEN_pdp_temp, MOX_pdp_temp, AOM_pdp_temp,MGEN_pdp_depth, MOX_pdp_depth,AOM_pdp_depth, ncol = 3, nrow = 3)

ggsave("Fig4.tiff", plot = Fig4, device = "tiff", dpi = 1000, width = 13, height = 11,  bg="white")

#supplement

#learning curves. Vertical
ggarrange(MGEN_LC,MOX_LC, AOM_LC, ncol = 1, nrow = 3)

```

