---
title: "Regression tree modeling: CH4 wetland fluxes"
author: "Satya Kent"
date: "2024-05-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load data and reorganize for regression tree modeling & GLM modeling
```{r}
library(tidyverse)

flux_rates=read_csv("flux_rates_meta-analysis.csv")

# Specify the order of x-axis categories
desired_order2 <- c("Methanogenesis", "Aerobic methane oxidation", "Anaerobic methane oxidation")

# Convert "Pathway" column to factor with desired order
flux_rates$Pathway <- factor(flux_rates$Pathway, levels = desired_order2)

#create filtered dataframe with all the environmental variables I care about.
flux_rates_filtered <- flux_rates[, c("Pathway", "Mean_rate", "Wetland_type", "Inc_temp", "Inc_length_days","Depth_representative", "Salinity_ppt", "Salinity_cat", "dom_veg", "method","subpathway")]


library(dplyr)

#keep NAs in categorical values instead of converting to Not Reported

#flux_rates_filtered2 <- flux_rates_filtered %>%
  #mutate(dom_veg = ifelse(is.na(dom_veg), "Not Reported", dom_veg),
         #subpathway = ifelse(is.na(subpathway), "Not Reported", subpathway),
         #Wetland_type = ifelse(is.na(Wetland_type), "Not Reported", Wetland_type),
         #Salinity_cat = ifelse(is.na(Salinity_cat), "Not Reported", Salinity_cat))

colnames(flux_rates_filtered)

# remove the continuous salinity column
flux_rates_filtered <- flux_rates_filtered %>%
  dplyr::select(-Salinity_cat)


#hot encoding of categorical values to integers (MLMs with a comb of categorical + numeric predictors)
# convert categorical variables to integers 
flux_rates_encoded <- flux_rates_filtered %>% mutate( Wetland_type = as.integer(factor(Wetland_type)), dom_veg = as.integer(factor(dom_veg)), method = as.integer(factor(method)), subpathway = as.integer(factor(subpathway)) ) 


#re-organize datasets for regression tree modeling 
MGEN_filtered=flux_rates_encoded[flux_rates_encoded$Pathway == "Methanogenesis", ]
MGEN_filtered$Pathway <- NULL
names(MGEN_filtered) <- c("Mean_rate", "Wetland_type", "Incubation_temperature", "Incubation_length", "Depth", "Salinity", "Dominant_vegetation", "Method", "Subpathway")

MOX_filtered=flux_rates_encoded[flux_rates_encoded$Pathway == "Aerobic methane oxidation", ]
MOX_filtered$Pathway <- NULL
names(MOX_filtered) <- c("Mean_rate", "Wetland_type", "Incubation_temperature", "Incubation_length", "Depth", "Salinity", "Dominant_vegetation", "Method", "Subpathway")
MOX_filtered$Subpathway <- NULL #no subpathways exist for MOx

AOM_filtered=flux_rates_encoded[flux_rates_encoded$Pathway == "Anaerobic methane oxidation", ]
AOM_filtered$Pathway <- NULL
names(AOM_filtered) <- c("Mean_rate", "Wetland_type", "Incubation_temperature", "Incubation_length", "Depth", "Salinity", "Dominant_vegetation", "Method", "Subpathway")
```



METHANOGENESIS

GLMs w gamma distribution & log link
```{r}
#GLMs do not assume normaliy and use a link function to allow for potentially non-linear relationships between the predictors and the actual response variable. Assumes observations are independent.


# Load necessary packages
library(dplyr)
library(broom)

# Prepare lists to store the results
results <- list()
coefficients_results <- list()

# Loop over each predictor variable
for (predictor in setdiff(colnames(MGEN_filtered), "Mean_rate")) {
  # Filter out NAs for the current predictor
  data_filtered <- MGEN_filtered %>%
    select(Mean_rate, all_of(predictor)) %>%
    drop_na()

  # Filter out rows where the predictor is "Not Reported" if it is a categorical variable
  if (is.factor(data_filtered[[predictor]]) || is.character(data_filtered[[predictor]])) {
    data_filtered <- data_filtered %>%
      filter(data_filtered[[predictor]] != "Not Reported")
  }

  # Fit the GLM using glm with Gamma family
  formula <- as.formula(paste("Mean_rate ~", predictor))
  glm_model <- glm(formula, data = data_filtered, family = Gamma(link = "log"))
  
  # Summarize the model using broom
  model_summary <- tidy(glm_model)
  deviance_explained <- 1 - (glm_model$deviance / glm_model$null.deviance)  # Pseudo R-squared
  p_value <- model_summary$p.value[2]  # P-value for the predictor
  AIC_value <- AIC(glm_model)  # AIC value for the model
  BIC_value <- BIC(glm_model)  # BIC value for the model
  sample_size <- nrow(data_filtered)  # Sample size after filtering

  # Store the results
  results[[predictor]] <- data.frame(
    Predictor = predictor,
    Pseudo_R_squared = deviance_explained,
    P_value = p_value,
    AIC = AIC_value,
    BIC = BIC_value
  )

  # Store coefficient results
  coefficients_results[[predictor]] <- data.frame(
    Predictor = predictor,
    Estimate = model_summary$estimate[2],
    Std_Error = model_summary$std.error[2],
    Z_value = model_summary$statistic[2],
    P_value = model_summary$p.value[2],
    Sample_Size = sample_size
  )
}

# Combine all results into single data frames
glm_results_df <- bind_rows(results)
coefficients_df <- bind_rows(coefficients_results)

# Sort by descending Pseudo R squared
glm_results_df <- glm_results_df %>%
  arrange(desc(as.numeric(glm_results_df$Pseudo_R_squared)))

# Bold significant Pseudo R squared values
glm_results_df$Pseudo_R_squared <- ifelse(as.numeric(glm_results_df$P_value) < 0.05, 
                                          paste0("**", round(as.numeric(glm_results_df$Pseudo_R_squared), 3), "**"), 
                                          round(as.numeric(glm_results_df$Pseudo_R_squared), 3))

# Print the results dataframes
print(glm_results_df)
print(coefficients_df)


```

GAMs w PDPs
```{r}
# Load necessary packages
library(dplyr)
library(mgcv)
library(gratia)  # Load gratia for plotting

# Prepare a list to store the results
results <- list()

# Loop over each predictor variable
for (predictor in setdiff(colnames(MGEN_filtered), "Mean_rate")) {
  # Filter out NAs for the current predictor
  data_filtered <- MGEN_filtered %>%
    select(Mean_rate, all_of(predictor)) %>%
    drop_na()

  # Filter out rows where the predictor is "Not Reported" if it is a categorical variable
  if (is.factor(data_filtered[[predictor]]) || is.character(data_filtered[[predictor]])) {
    data_filtered <- data_filtered %>%
      filter(data_filtered[[predictor]] != "Not Reported")
  }

  # Fit the GAM using gam from mgcv package
  formula <- if (is.numeric(data_filtered[[predictor]])) {
    as.formula(paste("Mean_rate ~ s(", predictor, ")", sep = ""))
  } else {
    as.formula(paste("Mean_rate ~", predictor))
  }
  
  gam_model <- gam(formula, data = data_filtered, family = Gamma(link = "log"))
  
  # Get the summary of the model
  gam_summary <- summary(gam_model)
  
  # Initialize p-value
  p_value <- NA
  
  # Extract p-value for the predictor
  if (is.numeric(data_filtered[[predictor]])) {
    # For smooth terms, use s.table
    if ("s.table" %in% names(gam_summary)) {
      s_table <- gam_summary$s.table
      if (nrow(s_table) > 0) {
        # Print s_table to check its structure
        print("Smooth Term Table:")
        print(s_table)
        # Extract p-value for the smooth term
        p_value <- s_table[1, "p-value"]  # Assuming single smooth term
      }
    }
  } else {
    # For parametric terms, use p.table
    if ("p.table" %in% names(gam_summary)) {
      p_table <- gam_summary$p.table
      # Print p_table to check its structure and row names
      print("Parametric Term Table:")
      print(p_table)
      if (predictor %in% rownames(p_table)) {
        p_value <- p_table[predictor, "Pr(>|t|)"]
      } else {
        # Check if predictor is in any level
        levels <- rownames(p_table)
        print(paste("Levels in p.table:", paste(levels, collapse = ", ")))
        p_value <- p_table[grep(predictor, levels), "Pr(>|t|)"]
      }
    }
  }

  # Extract R-squared from the summary
  r_squared <- gam_summary$r.sq
  
  # Calculate AIC and BIC
  AIC_value <- AIC(gam_model)
  BIC_value <- BIC(gam_model)

  # Store the results
  results[[predictor]] <- data.frame(
    Predictor = predictor,
    R_squared = r_squared,
    P_value = ifelse(is.numeric(p_value), mean(p_value, na.rm = TRUE), NA),  # If multiple p-values, take the mean
    AIC = AIC_value,
    BIC = BIC_value
  )
  
  # Create Partial Dependence Plots for continuous predictors
  if (is.numeric(data_filtered[[predictor]])) {
    # Plot smooth term for continuous predictor
    p <- draw(gam_model, select = 1)  # Select the first smooth term
    print(p)
  }
}

# Combine all results into a single data frame
gam_results_df <- bind_rows(results)

# Sort by descending R squared
gam_results_df <- gam_results_df %>%
  arrange(desc(R_squared))

# Print the results data frame
print("GAM Results:")
print(gam_results_df)

```


Boosted regression tree w caret
```{r}

# from Elith 2008:
#"The BRT approach differs fundamentally from traditional regression methods that produce a single ‘best’ model, instead using the technique of boosting to combine large numbers of relatively simple tree models adaptively, to optimize predictive performance". 

#"The final BRT model can be understood as an additive regression model in which individual terms are simple trees, fitted in a forward, stagewise fashion."


# Load necessary libraries
library(gbm)
library(ggplot2)
library(caret)


#Random Search 
random_grid <- data.frame( n.trees = sample(seq(500, 10000, by = 100), 15, replace = TRUE), interaction.depth = sample(seq(1, 10), 15, replace = TRUE), shrinkage = runif(15, 0.0001, 0.1), n.minobsinnode = sample(seq(1, 20), 15, replace = TRUE))

# Grid of hyperparameters
param_grid <- expand.grid(
  n.trees = c(500, 5000, 7000),  # Number of trees
  interaction.depth = c(1, 3, 5),  # Depth of trees
  shrinkage = c(0.001, 0.1),  # Learning rate
  n.minobsinnode = c(5, 10)  # Minimum observations in node
)

# Initialize a vector to store results
cv_errors <- numeric(nrow(random_grid))

# Iterate over the parameter grid
for (i in 1:nrow(random_grid)) {
  params <- random_grid[i, ]
 
  # Fit the model for the given parameters
  gbm_model <- gbm(
    formula = Mean_rate ~ Wetland_type+Incubation_temperature+Incubation_length+Depth+Salinity+Dominant_vegetation+Method+Subpathway,
    distribution = "gaussian",
    data = MGEN_filtered,
    n.trees = params$n.trees,
    interaction.depth = params$interaction.depth,
    shrinkage = params$shrinkage,
    n.minobsinnode = params$n.minobsinnode,
    cv.folds = 10,
    verbose = FALSE
  )
 
  # Get cross-validation error
  cv_errors[i] <- gbm.perf(gbm_model, method = "cv", plot.it = FALSE)
}

# Find the best parameter combination
best_params <- random_grid[which.min(cv_errors), ]
print(best_params)







# Set seed for reproducibility
set.seed(123)

#change categorical column classes to factors
MGEN_filtered <- MGEN_filtered %>%
  mutate_if(is.character, as.factor)

# Replace spaces with underscores in column names for future plotting
colnames(MGEN_filtered) <- gsub(" ", "_", colnames(MGEN_filtered))

# Shuffle rows in the dataframe
set.seed(123)
MGEN_filtered <- MGEN_filtered %>%
  sample_n(nrow(.))  # Randomly shuffle rows


# Set variables for model
data <- MGEN_filtered
response <- data$Mean_rate
predictors <- data %>% select(-Mean_rate)


# Perform a 5-fold cross-validation to find the optimal number of trees (recommended for small datasets) that minimizes RMSE. Recommended method by Ridgeway 2006 (author of the gbm package).'I usually aim for 3,000 to 10,000 iterations with shrinkage rates between 0.01 and 0.001"
#Further tune hyper-parameters according to reccommended by to minimize RMSE.

#Steps
#The dataset is split into k folds.

cv_folds <- 10

set.seed(123)
gbm_model_cv <- gbm(
  formula = Mean_rate ~ .,
  data = data,
  distribution = "gaussian", #for numerical data 
  n.trees = best_params$n.trees, #max x value on plot, change depending on best # trees identified
  interaction.depth = best_params$interaction.depth, #Elith 2008 recommends values of 2 or 3 for small sample sizes (<250)
  shrinkage = best_params$shrinkage, #what papers recommend for small sample sizes (Elith 2008; Ridgeway 2024), 
  #smaller values increase model performance 
  n.minobsinnode = best_params$n.minobsinnode, #default is 10
  cv.folds = cv_folds, #10-fold cv
  verbose = TRUE
)


# get MSE and compute RMSE. RSME = difference between values predicted by a model and the actual values.

sqrt(min(gbm_model_cv$cv.error))
## ~1.49

#Plot shows the cross-validation performance of the boosted regression tree model, showcasing the mean squared error on both training (black) and validation (green) datasets over boosting iterations. The plot helps identify the optimal number of trees to prevent overfitting while maximizing predictive accuracy.

best_ntrees <- gbm.perf(gbm_model_cv, method = "cv", plot.it = TRUE) #here the best # of trees is 5440

print(best_ntrees)

# compute R-squared value
cv_predictions <- predict(gbm_model_cv, data, n.trees = 5440)

# Calculate Total Sum of Squares (TSS)
tss <- sum((data$Mean_rate - mean(data$Mean_rate))^2)

# Calculate Residual Sum of Squares (RSS)
rss <- sum((data$Mean_rate - cv_predictions)^2)

# Calculate R-squared
r_squared <- 1 - (rss / tss)
print(paste("R-squared: ", r_squared))
#0.69

#let's see how the RMSE compares to the standard deviation of the dataset. It should be lower, which indicates the model is demonstrating some ability to learn.

sd(MGEN_filtered$Mean_rate)
# 1.88




# Train the final model using the optimal number of trees and all data
final_model <- gbm(
  formula = Mean_rate ~ .,
  data = data,
  distribution = "gaussian",
  n.trees = 5440,
  interaction.depth = 3,
  shrinkage = 0.001,
  n.minobsinnode = 10,
  verbose = TRUE
)


# Summary of the final model
print(summary(final_model))

# Calculate AIC
n <- nrow(data)  # Number of observations
k <- length(final_model$var.levels) + 1  # Number of parameters (number of predictors + 1 for the intercept)
log_likelihood <- -n/2 * log(2 * pi) - n/2 * log(rss/n) - n/2  # Gaussian log-likelihood
aic <- 2 * k - 2 * log_likelihood  # AIC formula
print(paste("AIC: ", aic))

#286.12

#Variable importance

##1.) Relative Importance

#Each feature in the dataset is assigned a relative importance score. These scores (on the y-axis) are measures of how much each feature contributes to the model's predictive accuracy, normalized so their sum equals 100. The higher the score, the more important the feature is in predicting the target variable. E.g. Dominant Vegetation is the most influential predictor, contributing __% to the model's accuracy.

#"At each split in each tree, gbm computes the improvement in the split-criterion (MSE for regression). gbm then averages the improvement made by each variable across all the trees that the variable is used. The variables with the largest average decrease in MSE are considered most important.


# Extract feature importance
importance <- summary(final_model, plotit = FALSE)

importance_df <- data.frame(
  Feature = importance$var,
  Importance = importance$rel.inf
)

importance_df <- importance_df[order(-importance_df$Importance),]

# Plot feature importance
ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Feature Importance: Methanogenesis",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Depth", "Dominant vegetation", "Incubation length","Subpathway", "Method", "Wetland type","Incubation temperature", "Salinity")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#FINAL FIG: Rainbow
rainbowMGEN=ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Methanogenesis",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Depth", "Dominant vegetation", "Incubation length","Subpathway", "Method", "Wetland type","Incubation temperature", "Salinity")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#377eb8",
                               "Subpathway" = "#984ea3", 
                               "Method"="#f781bf",
                               "Wetland_type"="#e6ab02",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#ff7f00")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = c("#a65628", "#4daf4a", "#377eb8", "#984ea3", "#f781bf", "#e6ab02","#e41a1c", "#ff7f00" )), 
        legend.position = "none") 

#FINAL FIG: depth, dom veg, temp colorized
triMGEN=ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Methanogenesis",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Depth", "Dominant vegetation", "Incubation length","Subpathway", "Method", "Wetland type","Incubation temperature", "Salinity")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Subpathway" = "#B3B3B3", 
                               "Method"="#B3B3B3",
                               "Wetland_type"="#B3B3B3",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = c("#a65628", "#4daf4a", "#999999", "#999999", "#999999", "#999999","#e41a1c", "#999999")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none")

#standardize order of features between plots based on sorted average importance across 3 pathways 
triMGEN_sorted=ggplot(importance_df, aes(x = factor(Feature, levels = c("Dominant_vegetation", 
                                                         "Incubation_temperature", 
                                                         "Depth", "Incubation_length", 
                                                         "Subpathway", "Method", 
                                                         "Wetland_type", "Salinity")), 
                          y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Methanogenesis",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Dominant vegetation", "Incubation temperature", "Depth", 
                              "Incubation length", "Subpathway", "Method", 
                              "Wetland type", "Salinity")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Subpathway" = "#B3B3B3", 
                               "Method" = "#B3B3B3",
                               "Wetland_type" = "#B3B3B3",
                               "Incubation_temperature" = "#e41a1c", 
                               "Salinity" = "#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, 
                                   color = c("#4daf4a", "#e41a1c", "#a65628", 
                                             "#B3B3B3", "#B3B3B3", "#B3B3B3", 
                                             "#B3B3B3", "#B3B3B3")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none")


# Normalize importance scores
total_importance <- sum(importance_df$Importance)
importance_df$Normalized_Importance <- importance_df$Importance / total_importance

# Plot the normalized importance scores
ggplot(importance_df, aes(x = reorder(Feature, -Normalized_Importance), y = Normalized_Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Feature Importance: MGEN",
       x = NULL,
       y = "Relative Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

##2.) Permutation Importance

# computes the Mean Squared Error (MSE) of the model predictions on the original dataset to those of the permuted dataset.


#Permutation importance scores are calculated by evaluating the change in model performance when each feature's values are randomly shuffled. Initially, the model's performance (Mean Squared Error, MSE) is measured on the original data. Then, for each feature, the values are permuted, and the model's performance is re-evaluated on this permuted data. The importance score for a feature is the difference between the permuted performance and the original performance. A larger positive difference indicates higher importance, as it shows that permuting the feature significantly degrades model performance.
#Features with higher values are crucial for predictions; permuting them significantly increases error (RSME).

# Initial model performance evaluation
original_perf <- with(MGEN_filtered, mean((Mean_rate - predict(final_model, newdata = MGEN_filtered, n.trees = 5440))^2))

# Storage for importance scores
feature_importance <- setNames(numeric(ncol(MGEN_filtered) - 1), names(MGEN_filtered)[-which(names(MGEN_filtered) == "Mean_rate")])

# Calculate importance for each feature
set.seed(123)  # For reproducibility
for (feature in names(feature_importance)) {
  # Copy the original data
  data_permuted <- MGEN_filtered

  # Permute the feature column
  data_permuted[[feature]] <- sample(data_permuted[[feature]])

  # Calculate performance with permuted data
  permuted_perf <- with(data_permuted, mean((Mean_rate - predict(final_model, newdata = data_permuted, n.trees = 5440))^2))

  # Importance score is the loss in performance
  feature_importance[feature] <- permuted_perf - original_perf
}

# Sorting features by importance
feature_importance <- sort(feature_importance, decreasing = TRUE)

# Print the importance
print(feature_importance)

# Convert to data frame for ggplot
importance_df <- data.frame(Feature = names(feature_importance), Importance = feature_importance)

# Plot
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Permutation Importance: Methanogenesis", x = NULL, y = "Decrease in Performance (MSE)") +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))


# permutation importance plot with color
permMGEN=ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Methanogenesis",
       x = NULL,
       y = "Decrease in Performance (MSE)") +
  scale_x_discrete(labels = c("Salinity","Incubation temperature","Wetland type","Method","Subpathway", "Incubation length","Dominant vegetation","Depth" )) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                                "Dominant_vegetation" = "#4daf4a", 
                                "Incubation_length" = "#B3B3B3",
                                "Subpathway" = "#B3B3B3", 
                                "Method" = "#B3B3B3",
                                "Wetland_type" = "#B3B3B3",
                                "Incubation_temperature" = "#e41a1c", 
                                "Salinity" = "#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1, 
                                   color = c( "#999999", "#e41a1c","#999999", "#999999", "#999999",  "#999999",  "#4daf4a","#a65628")), 
        plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")


## Partial dependency plots. 

#They visualize the marginal effect of each predictor variable on the predicted outcome, averaging out the effects of other predictors. The y axis represents how the Mean_rate is changing with the change in the given predictor variable.

library(pdp)
library(ggplot2)

for (predictor in setdiff(colnames(MGEN_filtered), "Mean_rate")) {
  partial_plot <- partial(final_model, pred.var = predictor, n.trees = 5440)
  
  # Convert partial_plot to data frame
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Determine the type of predictor
  predictor_type <- ifelse(is.numeric(MGEN_filtered[[predictor]]), "continuous", "categorical")
  
  # Filter out "Not Reported" data points for categorical predictors
  if (predictor_type == "categorical") {
    partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  }
  
  # Create ggplot based on predictor type
  if (predictor_type == "continuous") {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_line() +
      geom_rug(data = MGEN_filtered, aes(x = !!sym(predictor), y = NULL), sides = "t", color = "blue", size = 0.3) +
      labs(title = paste("Methanogenesis"),
           x = predictor, y = "Predicted rate") +
      theme_minimal()
  } else {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_point(size = 15, shape = 95) +  
      labs(title = paste("Methanogenesis"),
           x = predictor, y = "Predicted rate") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  }
  
  # Print the plot
  print(p)
}

#save individual pdp's (dom_veg, depth, inc_temp) for the paper to be plotted 

library(pdp)
library(ggplot2)

# Generate and store partial dependency plot for Depth
depth_MGEN <- {
  partial_plot <- partial(final_model, pred.var = "Depth", n.trees = 5440)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")

  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_line() +
    geom_rug(data = MGEN_filtered, aes(x = Depth, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Depth (cm)", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal()
}

# Generate and store partial dependency plot for Dominant Vegetation
dom_veg_MGEN <- {
  partial_plot <- partial(final_model, pred.var = "Dominant_vegetation", n.trees = 5440)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Filter out "Not Reported" data points
  partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  
  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_point(size = 8, shape = 95) +  
    labs(title = "Methanogenesis",
         x = "Dominant Vegetation", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))
}

# Generate and store partial dependency plot for Dominant Vegetation. Sort x axis categories from lowest to highest predicted rate
dom_veg_MGEN_sorted <- {
  partial_plot <- partial(final_model, pred.var = "Dominant_vegetation", n.trees = 5440)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Filter out "Not Reported" data points
  partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  
  ggplot(partial_df, aes(x = reorder(Value, Response), y = Response)) +  # Reorder based on Response
    geom_point(size = 8, shape = 95) +  
    labs(title = "Methanogenesis",
         x = "Dominant Vegetation", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))
}


# Generate and store partial dependency plot for Incubation Temperature
incub_temp_MGEN <- {
  partial_plot <- partial(final_model, pred.var = "Incubation_temperature", n.trees = 5440)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")

  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_line() +
    geom_rug(data = MGEN_filtered, aes(x = Incubation_temperature, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Incubation Temperature (\u00b0C)", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal()
}


```


Univariate regression tree
```{r}

#packages for regression trees
library(rpart)
library(rpart.plot)

set.seed(123)
tree_univariate_mgen <- rpart(Mean_rate ~. , data = MGEN_filtered, method = "anova")

#visualize the tree
rpart.plot(tree_univariate_mgen)

#let's check the residuals
plot(predict(tree_univariate_mgen),residuals(tree_univariate_mgen))

## Pruning the tree

#Goal: see if a smaller subtree can give us comparable results to the fully grown tree. If yes, we should go for the simpler tree because it reduces the likelihood of overfitting.

#A robust strategy of pruning the tree consists of avoiding splitting a partition if the split does not significantly improve the overall quality of the model.

#In the rpart package, this is controlled by the complexity parameter (cp), which imposes a penalty to the tree for having two many splits. The default value in rpart() is 0.01. The higher the cp, the smaller the tree.

#A too small value of cp leads to overfitting and a too large cp value will result in a too small tree. Both cases decrease the predictive performance of the model.

#An optimal cp value can be estimated by testing different cp values and using cross-validation approaches to determine the corresponding prediction accuracy of the model. The best cp is then defined as the one that maximizes the cross-validation accuracy.

# In the plot generated by plotcp(), "A good choice of cp for pruning is often the leftmost value for which the mean lies below the horizontal line"
# Here, this is cp = 0.093

plotcp(tree_univariate_mgen)

#Prints a table of optimal prunings based on a complexity parameter
printcp(tree_univariate_mgen)

#visualize more levels
tree_univariate_mgen2 <- rpart(Mean_rate ~. , data = MGEN_filtered, method = "anova", control=list(cp=0,xval=10))

plotcp(tree_univariate_mgen2)

printcp(tree_univariate_mgen2)

# Prune the model based on the optimal cp value
tree_uni_mgen_pruned <- prune(tree_univariate_mgen, cp = 0.093)

rpart.plot(tree_uni_mgen_pruned)

printcp(tree_uni_mgen_pruned)


#let's check the residuals
plot(predict(tree_uni_mgen_pruned),residuals(tree_uni_mgen_pruned))


#Extract importance values from tree
importance = tree_uni_mgen_pruned$variable.importance

importance_df = data.frame(
  Feature = names(importance),
  Importance = importance
)

importance_df = importance_df[order(-importance_df$Importance),]

#Gini importance 
ggplot(importance_df, aes(x=reorder(Feature, -Importance), y=Importance))+
  geom_col()+
  theme_minimal()+
  labs(title="Feature Importance: Methanogenesis ",
       x="Features", 
       y= "Gini Importance")+
  theme(axis.text.x = element_text(angle = 45, hjust =1))

## Calculate the sum of importance scores
total_importance <- sum(importance_df$Importance)

# Normalize importance scores
importance_df$Normalized_Importance <- importance_df$Importance / total_importance

# Plot the normalized importance scores
ggplot(importance_df, aes(x=reorder(Feature, -Normalized_Importance), y=Normalized_Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title="Relative Feature Importance: MGEN",
       x = NULL, 
       y= "Gini Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust =1))


#Permutation importance 

# Initial model performance evaluation
original_perf <- with(MGEN_filtered, mean((Mean_rate - predict(tree_uni_mgen_pruned, newdata = MGEN_filtered))^2))

# Storage for importance scores
feature_importance <- setNames(numeric(ncol(MGEN_filtered) - 1), names(MGEN_filtered)[-which(names(MGEN_filtered) == "Mean_rate")])

# Calculate importance for each feature
set.seed(123)  # For reproducibility
for (feature in names(feature_importance)) {
  # Copy the original data
  data_permuted <- MGEN_filtered
 
  # Permute the feature column
  data_permuted[[feature]] <- sample(data_permuted[[feature]])
 
  # Calculate performance with permuted data
  permuted_perf <- with(data_permuted, mean((Mean_rate - predict(tree_uni_mgen_pruned, newdata = data_permuted))^2))
 
  # Importance score is the loss in performance
  feature_importance[feature] <- permuted_perf - original_perf
}

# Sorting features by importance
feature_importance <- sort(feature_importance, decreasing = TRUE)

# Print the importance
print(feature_importance)

# Convert to data frame for ggplot
importance_df <- data.frame(Feature = names(feature_importance), Importance = feature_importance)

# Plot
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Permutation Importance: Methanogenesis", x = NULL, y = "Decrease in Performance (MSE)") +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))


##Partial dependency plots
library(pdp)
library(ggplot2)

for (predictor in setdiff(colnames(MGEN_filtered), "Mean_rate")) {
  partial_plot <- partial(tree_uni_mgen_pruned, pred.var = predictor)
  
  # Convert partial_plot to data frame
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Determine the type of predictor
  predictor_type <- ifelse(is.numeric(MGEN_filtered[[predictor]]), "continuous", "categorical")
  
  # Filter out "Not Reported" data points for categorical predictors. We don't want to plot these.
  if (predictor_type == "categorical") {
    partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  }
  
  # Create ggplot based on predictor type
  if (predictor_type == "continuous") {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_line() +
      geom_point() +  
      labs(title = paste("Partial Dependency Plot for", predictor),
           x = predictor, y = "Predicted rate") +
      theme_minimal()
  } else {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_point(size = 15, shape = 95) +  
      labs(title = paste("Partial Dependency Plot for", predictor),
           x = predictor, y = "Predicted rate") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  }
  
  # Print the plot
  print(p)
}


```

Shapley values from pruned univariate regression tree
```{r}

# Load necessary libraries
library(kernelshap)
library(shapviz)
library(ggplot2)


# Set seed for reproducibility
set.seed(123)

# Define the full dataset for both rows to be explained and background data. I'm doing this since my dataset is so small (n=93).

X <- MGEN_filtered[, -which(names(MGEN_filtered) == "Mean_rate")]  # Exclude target variable
bg_X <- X  # Use the same dataset for background data

# Define a prediction function for the model
pred_fn <- function(model, newdata) {
  predict(model, newdata = newdata)
}

# Calculate SHAP values using kernelshap
shap_values <- kernelshap(tree_uni_mgen_pruned, X = X, bg_X = bg_X, pred_wrapper = pred_fn, nsim = 100)

# Convert SHAP values to a shapviz object for analysis
shap_viz <- shapviz(shap_values)

# Analyze and visualize SHAP values

# Plot feature importance
sv_importance(shap_viz) #looks similar to permutation importance plot
sv_importance(shap_viz, kind = "bee") # plot that shows directionality is not well suited to my dataset bc the most important predictors (dominant vegetation, wetland type) are categorical.  

```


AEROBIC METHANE OXIDATION

GLMs w gamma distribution & log link
```{r}
#GLMs do not assume normaliy and use a link function to allow for potentially non-linear relationships between the predictors and the actual response variable. Assumes observations are independent.


#Note: The predictor Salinity is removed from this model because it only has one level (low) for this pathway and it needs 2 to run. 

# Load necessary packages
library(dplyr)
library(broom)

# Prepare lists to store the results
results <- list()
coefficients_results <- list()

# Loop over each predictor variable
for (predictor in setdiff(colnames(MOX_filtered), c("Mean_rate", "Salinity"))) {  # Exclude Salinity from analysis
  # Filter out NAs for the current predictor
  data_filtered <- MOX_filtered %>%
    select(Mean_rate, all_of(predictor)) %>%
    drop_na()

  # Filter out rows where the predictor is "Not Reported" if it is a categorical variable
  if (is.factor(data_filtered[[predictor]]) || is.character(data_filtered[[predictor]])) {
    data_filtered <- data_filtered %>%
      filter(data_filtered[[predictor]] != "Not Reported")
  }

  # Fit the GLM using glm with Gamma family
  formula <- as.formula(paste("Mean_rate ~", predictor))
  glm_model <- glm(formula, data = data_filtered, family = Gamma(link = "log"))
  
  # Summarize the model using broom
  model_summary <- tidy(glm_model)
  deviance_explained <- 1 - (glm_model$deviance / glm_model$null.deviance)  # Pseudo R-squared
  p_value <- model_summary$p.value[2]  # P-value for the predictor
  AIC_value <- AIC(glm_model)  # AIC value for the model
  BIC_value <- BIC(glm_model)  # BIC value for the model
  sample_size <- nrow(data_filtered)  # Sample size after filtering

  # Store the results
  results[[predictor]] <- data.frame(
    Predictor = predictor,
    Pseudo_R_squared = deviance_explained,
    P_value = p_value,
    AIC = AIC_value,
    BIC = BIC_value
  )

  # Store coefficient results
  coefficients_results[[predictor]] <- data.frame(
    Predictor = predictor,
    Estimate = model_summary$estimate[2],
    Std_Error = model_summary$std.error[2],
    Z_value = model_summary$statistic[2],
    P_value = model_summary$p.value[2],
    Sample_Size = sample_size
  )
}

# Combine all results into single data frames
glm_results_df <- bind_rows(results)
coefficients_df <- bind_rows(coefficients_results)

# Sort by descending Pseudo R squared
glm_results_df <- glm_results_df %>%
  arrange(desc(as.numeric(glm_results_df$Pseudo_R_squared)))

# Bold significant Pseudo R squared values
glm_results_df$Pseudo_R_squared <- ifelse(as.numeric(glm_results_df$P_value) < 0.05, 
                                          paste0("**", round(as.numeric(glm_results_df$Pseudo_R_squared), 3), "**"), 
                                          round(as.numeric(glm_results_df$Pseudo_R_squared), 3))

# Print the results dataframes
print(glm_results_df)
print(coefficients_df)



```

Boosted regression tree
(Failed to converge due to small sample size so decreased the minimum number of observations in the terminal nodes from 10 to 3)
```{r}

# from Elith 2008:
#"The BRT approach differs fundamentally from traditional regression methods that produce a single ‘best’ model, instead using the technique of boosting to combine large numbers of relatively simple tree models adaptively, to optimize predictive performance". 

#"The final BRT model can be understood as an additive regression model in which individual terms are simple trees, fitted in a forward, stagewise fashion."


# Load necessary libraries
library(gbm)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

#change categorical column classes to factors
MOX_filtered <- MOX_filtered %>%
  mutate_if(is.character, as.factor)

# Replace spaces with underscores in column names for future plotting
colnames(MOX_filtered) <- gsub(" ", "_", colnames(MOX_filtered))


# Shuffle rows in the dataframe
set.seed(123)
MOX_filtered <- MOX_filtered %>%
  sample_n(nrow(.))  # Randomly shuffle rows

# Set variables for model
data <- MOX_filtered
response <- data$Mean_rate
predictors <- data %>% select(-Mean_rate)


# Perform a 5-fold cross-validation to find the optimal number of trees (recommended for small datasets). Recommended method by Ridgeway 2024 (author of the gbm package).

#Steps
#The dataset is split into k folds.

cv_folds <- 5

set.seed(123)
gbm_model_cv <- gbm(
  formula = Mean_rate ~ .,
  data = data,
  distribution = "gaussian", #for numerical data 
  n.trees = 5000, #max x value on plot
  interaction.depth = 3, #Elith 2008 recommends values of 2 or 3 for small sample sizes (<250)
  shrinkage = 0.001, #what papers recommend for small sample sizes (Elith 2008; Ridgeway 2024), 
  #smaller values increase model performance 
  n.minobsinnode = 3, #reduced here from 10 since the sample size is so small 
  cv.folds = cv_folds, #5-fold cv
  verbose = TRUE
)

# get MSE and compute RMSE. RSME = difference between values predicted by a model and the actual values.

sqrt(min(gbm_model_cv$cv.error))
## 2.97


#Plot shows the cross-validation performance of the boosted regression tree model, showcasing the mean squared error on both training (black) and validation (green) datasets over boosting iterations. The plot helps identify the optimal number of trees to prevent overfitting while maximizing predictive accuracy.

best_ntrees <- gbm.perf(gbm_model_cv, method = "cv", plot.it = TRUE) #here the best # of trees is 4554
print(best_ntrees)

# compute R-squared value
cv_predictions <- predict(gbm_model_cv, data, n.trees = 4554)

# Calculate Total Sum of Squares (TSS)
tss <- sum((data$Mean_rate - mean(data$Mean_rate))^2)


# Calculate Residual Sum of Squares (RSS)
rss <- sum((data$Mean_rate - cv_predictions)^2)

# Calculate R-squared
r_squared <- 1 - (rss / tss)
print(paste("R-squared: ", r_squared))
#0.65

#let's see how the RMSE compares to the standard deviation of the dataset. It should be lower, which indicates the model is demonstrating some ability to learn.

sd(MOX_filtered$Mean_rate)
# 3.25



# Train the final model using the optimal number of trees and all data
final_model <- gbm(
  formula = Mean_rate ~ .,
  data = data,
  distribution = "gaussian",
  n.trees = 4554,
  interaction.depth = 3,
  shrinkage = 0.001,
  n.minobsinnode = 3,
  verbose = TRUE
)


# Summary of the final model
print(summary(final_model))


# Calculate AIC
n <- nrow(data)  # Number of observations
k <- length(final_model$var.levels) + 1  # Number of parameters (number of predictors + 1 for the intercept)
log_likelihood <- -n/2 * log(2 * pi) - n/2 * log(rss/n) - n/2  # Gaussian log-likelihood
aic <- 2 * k - 2 * log_likelihood  # AIC formula
print(paste("AIC: ", aic))

#105.87

#Variable importance

##1.) Relative Importance

#Each feature in the dataset is assigned a relative importance score. These scores (on the y-axis) are measures of how much each feature contributes to the model's predictive accuracy, normalized so their sum equals 100. The higher the score, the more important the feature is in predicting the target variable. E.g. Dominant Vegetation is the most influential predictor, contributing __% to the model's accuracy.

#"At each split in each tree, gbm computes the improvement in the split-criterion (MSE for regression). gbm then averages the improvement made by each variable across all the trees that the variable is used. The variables with the largest average decrease in MSE are considered most important.


# Extract feature importance
importance <- summary(final_model, plotit = FALSE)

importance_df <- data.frame(
  Feature = importance$var,
  Importance = importance$rel.inf
)

importance_df <- importance_df[order(-importance_df$Importance),]

# Plot feature importance
ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Feature Importance: MOx",
       x = "Features",
       y = "Relative Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#FINAL FIG: Rainbow
rainbowMOx=ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Aerobic methane oxidation",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Dominant vegetation","Wetland type","Method","Depth","Incubation length","Incubation temperature", "Salinity")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#377eb8",
                               "Method"="#f781bf",
                               "Wetland_type"="#e6ab02",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#ff7f00")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = c("#4daf4a", "#e6ab02", "#f781bf", "#a65628", "#377eb8","#e41a1c", "#ff7f00" )), 
        legend.position = "none") 

#FINAL FIG: depth, dom veg, temp highlighted
triMOx=ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Aerobic methane oxidation",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Dominant vegetation","Wetland type","Method","Depth","Incubation length","Incubation temperature", "Salinity")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Method"="#B3B3B3",
                               "Wetland_type"="#B3B3B3",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = c("#4daf4a", "#999999", "#999999", "#a65628", "#999999","#e41a1c", "#999999")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none")

#standardize order of features between plots based on sorted average importance across 3 pathways 
triMOx_sorted=ggplot(importance_df, aes(x = factor(Feature, levels = c("Dominant_vegetation", 
                                                         "Incubation_temperature", 
                                                         "Depth", "Incubation_length", 
                                                       "Method", 
                                                         "Wetland_type", "Salinity")), 
                          y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Aerobic methane oxidation",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Dominant vegetation", "Incubation temperature", "Depth", 
                              "Incubation length", "Method", 
                              "Wetland type", "Salinity")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Method" = "#B3B3B3",
                               "Wetland_type" = "#B3B3B3",
                               "Incubation_temperature" = "#e41a1c", 
                               "Salinity" = "#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, 
                                   color = c("#4daf4a", "#e41a1c", "#a65628", 
                                             "#B3B3B3", "#B3B3B3", "#B3B3B3", 
                                             "#B3B3B3")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none")

# Normalize importance scores
total_importance <- sum(importance_df$Importance)
importance_df$Normalized_Importance <- importance_df$Importance / total_importance

# Plot the normalized importance scores
ggplot(importance_df, aes(x = reorder(Feature, -Normalized_Importance), y = Normalized_Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Feature Importance: MOx",
       x = NULL,
       y = "Relative Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

##2.) Permutation Importance

# computes the Mean Squared Error (MSE) of the model predictions on the original dataset to those of the permuted dataset.


#Permutation importance scores are calculated by evaluating the change in model performance when each feature's values are randomly shuffled. Initially, the model's performance (Mean Squared Error, MSE) is measured on the original data. Then, for each feature, the values are permuted, and the model's performance is re-evaluated on this permuted data. The importance score for a feature is the difference between the permuted performance and the original performance. A larger positive difference indicates higher importance, as it shows that permuting the feature significantly degrades model performance.
#Features with higher values are crucial for predictions; permuting them significantly increases error (RSME).

# Initial model performance evaluation
original_perf <- with(MOX_filtered, mean((Mean_rate - predict(final_model, newdata = MOX_filtered, n.trees = 4554))^2))

# Storage for importance scores
feature_importance <- setNames(numeric(ncol(MOX_filtered) - 1), names(MOX_filtered)[-which(names(MOX_filtered) == "Mean_rate")])

# Calculate importance for each feature
set.seed(123)  # For reproducibility
for (feature in names(feature_importance)) {
  # Copy the original data
  data_permuted <- MOX_filtered

  # Permute the feature column
  data_permuted[[feature]] <- sample(data_permuted[[feature]])

  # Calculate performance with permuted data
  permuted_perf <- with(data_permuted, mean((Mean_rate - predict(final_model, newdata = data_permuted, n.trees = 4554))^2))

  # Importance score is the loss in performance
  feature_importance[feature] <- permuted_perf - original_perf
}

# Sorting features by importance
feature_importance <- sort(feature_importance, decreasing = TRUE)

# Print the importance
print(feature_importance)

# Convert to data frame for ggplot
importance_df <- data.frame(Feature = names(feature_importance), Importance = feature_importance)

# Plot
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Permutation Importance: MOx", x = NULL, y = "Decrease in Performance (MSE)") +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))

# permutation importance plot with color
permMOx=ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Aerobic methane oxidation",
       x = NULL,
       y = "Decrease in Performance (MSE)") +
  scale_x_discrete(labels = c("Salinity","Incubation temperature","Incubation length","Method","Wetland type","Depth", "Dominant vegetation" )) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                                "Dominant_vegetation" = "#4daf4a", 
                                "Incubation_length" = "#B3B3B3",
                                "Subpathway" = "#B3B3B3", 
                                "Method" = "#B3B3B3",
                                "Wetland_type" = "#B3B3B3",
                                "Incubation_temperature" = "#e41a1c", 
                                "Salinity" = "#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1, 
                                   color = c( "#999999", "#e41a1c","#999999", "#999999", "#999999", "#a65628", "#4daf4a")), 
        plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")


## Partial dependency plots. 

#They visualize the marginal effect of each predictor variable on the predicted outcome, averaging out the effects of other predictors. The y axis represents how the Mean_rate is changing with the change in the given predictor variable.



library(pdp)
library(ggplot2)

for (predictor in setdiff(colnames(MOX_filtered), "Mean_rate")) {
  partial_plot <- partial(final_model, pred.var = predictor, n.trees = 4554)
  
  # Convert partial_plot to data frame
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Determine the type of predictor
  predictor_type <- ifelse(is.numeric(MOX_filtered[[predictor]]), "continuous", "categorical")
  
  # Filter out "Not Reported" data points for categorical predictors
  if (predictor_type == "categorical") {
    partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  }
  
  # Create ggplot based on predictor type
  if (predictor_type == "continuous") {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_line() +
      geom_rug(data = MOX_filtered, aes(x = !!sym(predictor), y = NULL), sides = "t", color = "blue", size = 0.3) +
      labs(title = paste("Aerobic methane oxidation"),
           x = predictor, y = "Predicted rate") +
      theme_minimal()
  } else {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_point(size = 15, shape = 95) +  
      labs(title = paste("Aerobic methane oxidation"),
           x = predictor, y = "Predicted rate") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  }
  
  # Print the plot
  print(p)
}

#save individual pdp's (dom_veg, depth, inc_temp) for the paper to be plotted 

library(pdp)
library(ggplot2)

# Generate and store partial dependency plot for Depth
depth_MOx <- {
  partial_plot <- partial(final_model, pred.var = "Depth", n.trees = 4554)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")

  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_line() +
    geom_rug(data = MOX_filtered, aes(x = Depth, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Depth (cm)", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal()
}

# Generate and store partial dependency plot for Dominant Vegetation
dom_veg_MOx <- {
  partial_plot <- partial(final_model, pred.var = "Dominant_vegetation", n.trees = 4554)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Filter out "Not Reported" data points
  partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  
  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_point(size = 8, shape = 95) +  
    labs(title = "Aerobic methane oxidation",
         x = "Dominant Vegetation", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))
}

# Generate and store partial dependency plot for Dominant Vegetation. Sort x axis categories from lowest to highest predicted rate
dom_veg_MOx_sorted <- {
  partial_plot <- partial(final_model, pred.var = "Dominant_vegetation", n.trees = 4554)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Filter out "Not Reported" data points
  partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  
  ggplot(partial_df, aes(x = reorder(Value, Response), y = Response)) +  # Reorder based on Response
    geom_point(size = 8, shape = 95) +  
    labs(title = "Aerobic methane oxidation",
         x = "Dominant Vegetation", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))
}


# Generate and store partial dependency plot for Incubation Temperature
incub_temp_MOx <- {
  partial_plot <- partial(final_model, pred.var = "Incubation_temperature", n.trees = 4554)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")

  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_line() +
    geom_rug(data = MOX_filtered, aes(x = Incubation_temperature, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Incubation Temperature (\u00b0C)", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal()
}


```

Univariate regression tree
```{r}

#packages for regression trees
library(rpart)
library(rpart.plot)

set.seed(123)
tree_univariate_MOX <- rpart(Mean_rate ~. , data = MOX_filtered, method = "anova")

#visualize the tree
rpart.plot(tree_univariate_MOX)

#let's check the residuals
plot(predict(tree_univariate_MOX),residuals(tree_univariate_MOX))

## Pruning the tree

#"Our goal here is to see if a smaller subtree can give us comparable results to the fully grown tree. If yes, we should go for the simpler tree because it reduces the likelihood of overfitting.

#One possible robust strategy of pruning the tree (or stopping the tree to grow) consists of avoiding splitting a partition if the split does not significantly improves the overall quality of the model.

#In rpart package, this is controlled by the complexity parameter (cp), which imposes a penalty to the tree for having two many splits. The default value is 0.01. The higher the cp, the smaller the tree.

#A too small value of cp leads to overfitting and a too large cp value will result to a too small tree. Both cases decrease the predictive performance of the model.

#An optimal cp value can be estimated by testing different cp values and using cross-validation approaches to determine the corresponding prediction accuracy of the model. The best cp is then defined as the one that maximizes the cross-validation accuracy.

# In plotcp() "A good choice of cp for pruning is often the leftmost value for which the mean lies below the horizontal line"
# Here, this is cp = 0.057

plotcp(tree_univariate_MOX)

#Prints a table of optimal prunings based on a complexity parameter
printcp(tree_univariate_MOX)

#visualize more levels
tree_univariate_MOX2 <- rpart(Mean_rate ~. , data = MOX_filtered, method = "anova", control=list(cp=0,xval=10))

plotcp(tree_univariate_MOX2)

printcp(tree_univariate_MOX2)

# Prune the model based on the optimal cp value
tree_uni_MOX_pruned <- prune(tree_univariate_MOX, cp = 0.057)

rpart.plot(tree_uni_MOX_pruned)

printcp(tree_uni_MOX_pruned)


#let's check the residuals
plot(predict(tree_uni_MOX_pruned),residuals(tree_uni_MOX_pruned))


#Extract importance values from tree
importance = tree_uni_MOX_pruned$variable.importance

importance_df = data.frame(
  Feature = names(importance),
  Importance = importance
)

importance_df = importance_df[order(-importance_df$Importance),]

#Gini importance 
ggplot(importance_df, aes(x=reorder(Feature, -Importance), y=Importance))+
  geom_col()+
  theme_minimal()+
  labs(title="Feature Importance: MOx ",
       x="Features", 
       y= "Gini Importance")+
  theme(axis.text.x = element_text(angle = 45, hjust =1))

## Calculate the sum of importance scores
total_importance <- sum(importance_df$Importance)

# Normalize importance scores
importance_df$Normalized_Importance <- importance_df$Importance / total_importance

# Plot the normalized importance scores
ggplot(importance_df, aes(x=reorder(Feature, -Normalized_Importance), y=Normalized_Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title="Relative Feature Importance: MOx",
       x = NULL, 
       y= "Gini Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust =1))


#Permutation importance 

# Initial model performance evaluation
original_perf <- with(MOX_filtered, mean((Mean_rate - predict(tree_uni_MOX_pruned, newdata = MOX_filtered))^2))

# Storage for importance scores
feature_importance <- setNames(numeric(ncol(MOX_filtered) - 1), names(MOX_filtered)[-which(names(MOX_filtered) == "Mean_rate")])

# Calculate importance for each feature
set.seed(123)  # For reproducibility
for (feature in names(feature_importance)) {
  # Copy the original data
  data_permuted <- MOX_filtered
 
  # Permute the feature column
  data_permuted[[feature]] <- sample(data_permuted[[feature]])
 
  # Calculate performance with permuted data
  permuted_perf <- with(data_permuted, mean((Mean_rate - predict(tree_uni_MOX_pruned, newdata = data_permuted))^2))
 
  # Importance score is the loss in performance
  feature_importance[feature] <- permuted_perf - original_perf
}

# Sorting features by importance
feature_importance <- sort(feature_importance, decreasing = TRUE)

# Print the importance
print(feature_importance)

# Convert to data frame for ggplot
importance_df <- data.frame(Feature = names(feature_importance), Importance = feature_importance)

# Plot
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Permutation Importance: MOx", x = NULL, y = "Decrease in Performance (MSE)") +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))


##Partial dependency plots
library(pdp)
library(ggplot2)

for (predictor in setdiff(colnames(MOX_filtered), "Mean_rate")) {
  partial_plot <- partial(tree_uni_MOX_pruned, pred.var = predictor)
  
  # Convert partial_plot to data frame
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Determine the type of predictor
  predictor_type <- ifelse(is.numeric(MOX_filtered[[predictor]]), "continuous", "categorical")
  
  # Filter out "Not Reported" data points for categorical predictors. We don't want to plot these.
  if (predictor_type == "categorical") {
    partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  }
  
  # Create ggplot based on predictor type
  if (predictor_type == "continuous") {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_line() +
      geom_point() +  
      labs(title = paste("Partial Dependency Plot for", predictor),
           x = predictor, y = "Predicted rate") +
      theme_minimal()
  } else {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_point(size = 15, shape = 95) +  
      labs(title = paste("Partial Dependency Plot for", predictor),
           x = predictor, y = "Predicted rate") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  }
  
  # Print the plot
  print(p)
}

```

Shapley values from pruned univariate regression tree
```{r}

# Load necessary libraries
library(kernelshap)
library(shapviz)
library(ggplot2)


# Set seed for reproducibility
set.seed(123)

# Define the full dataset for both rows to be explained and background data. I'm doing this since my dataset is so small (n=93).

X <- MOX_filtered[, -which(names(MOX_filtered) == "Mean_rate")]  # Exclude target variable
bg_X <- X  # Use the same dataset for background data

# Define a prediction function for the model
pred_fn <- function(model, newdata) {
  predict(model, newdata = newdata)
}

# Calculate SHAP values using kernelshap
shap_values <- kernelshap(tree_uni_MOX_pruned, X = X, bg_X = bg_X, pred_wrapper = pred_fn, nsim = 100)

# Convert SHAP values to a shapviz object for analysis
shap_viz <- shapviz(shap_values)

# Analyze and visualize SHAP values

# Plot feature importance
sv_importance(shap_viz) #looks similar to permutation importance plot
sv_importance(shap_viz, kind = "bee") # plot that shows directionality is not well suited to my dataset bc the most important predictors (dominant vegetation, wetland type) are categorical.  

```

ANAEROBIC METHANE OXIDATION

GLMs w gamma distribution & log link
(Failed to converge with the Incubation length parameter so I removed it and increased max iterations up to 100.)
```{r}
#GLMs do not assume normality and use a link function to allow for potentially non-linear relationships between the predictors and the actual response variable. Assumes observations are independent.

# Load necessary packages
library(dplyr)
library(broom)

# Prepare lists to store the results
results <- list()
coefficients_results <- list()

# Loop over each predictor variable
for (predictor in setdiff(colnames(AOM_filtered), "Mean_rate")) {
  # Filter out NAs for the current predictor
  data_filtered <- AOM_filtered %>%
    select(Mean_rate, all_of(predictor)) %>%
    drop_na()

  # Filter out rows where the predictor is "Not Reported" if it is a categorical variable
  if (is.factor(data_filtered[[predictor]]) || is.character(data_filtered[[predictor]])) {
    data_filtered <- data_filtered %>%
      filter(data_filtered[[predictor]] != "Not Reported")
  }

  # Fit the GLM using glm with Gamma family
  tryCatch({
    formula <- as.formula(paste("Mean_rate ~", predictor))
    glm_model <- glm(formula, data = data_filtered, family = Gamma(link = "log"), control = glm.control(maxit = 100)) # Increase max iterations to 100 which helps model convergence

    # Summarize the model using broom
    model_summary <- tidy(glm_model)
    deviance_explained <- 1 - (glm_model$deviance / glm_model$null.deviance)  # Pseudo R-squared
    p_value <- model_summary$p.value[2]  # P-value for the predictor
    AIC_value <- AIC(glm_model)  # AIC value for the model
    BIC_value <- BIC(glm_model)  # BIC value for the model
    sample_size <- nrow(data_filtered)  # Sample size after filtering

    # Store the results
    results[[predictor]] <- data.frame(
      Predictor = predictor,
      Pseudo_R_squared = deviance_explained,
      P_value = p_value,
      AIC = AIC_value,
      BIC = BIC_value
    )

    # Store coefficient results
    coefficients_results[[predictor]] <- data.frame(
      Predictor = predictor,
      Estimate = model_summary$estimate[2],
      Std_Error = model_summary$std.error[2],
      Z_value = model_summary$statistic[2],
      P_value = model_summary$p.value[2],
      Sample_Size = sample_size
    )
  }, error = function(e) {
    # Print a message if an error occurs
    message("Error with predictor ", predictor, ": ", e$message)
  })
}

# Combine all results into single data frames
glm_results_df <- bind_rows(results)
coefficients_df <- bind_rows(coefficients_results)

# Sort by descending Pseudo R squared
glm_results_df <- glm_results_df %>%
  arrange(desc(as.numeric(glm_results_df$Pseudo_R_squared)))

# Bold significant Pseudo R squared values
glm_results_df$Pseudo_R_squared <- ifelse(as.numeric(glm_results_df$P_value) < 0.05, 
                                          paste0("**", round(as.numeric(glm_results_df$Pseudo_R_squared), 3), "**"), 
                                          round(as.numeric(glm_results_df$Pseudo_R_squared), 3))

# Print the results dataframes
print(glm_results_df)
print(coefficients_df)

```


Boosted regression tree
```{r}

# from Elith 2008:
#"The BRT approach differs fundamentally from traditional regression methods that produce a single ‘best’ model, instead using the technique of boosting to combine large numbers of relatively simple tree models adaptively, to optimize predictive performance". 

#"The final BRT model can be understood as an additive regression model in which individual terms are simple trees, fitted in a forward, stagewise fashion."


# Load necessary libraries
library(gbm)
library(ggplot2)
library(dplyr)
library(pdp)

# Set seed for reproducibility
set.seed(123)

#change categorical column classes to factors
AOM_filtered <- AOM_filtered %>%
  mutate_if(is.character, as.factor)

# Replace spaces with underscores in column names for future plotting
colnames(AOM_filtered) <- gsub(" ", "_", colnames(AOM_filtered))

# Shuffle rows in the dataframe
set.seed(123)  # Ensure reproducibility of row shuffling
AOM_filtered <- AOM_filtered %>%
  sample_n(nrow(.))  # Randomly shuffle rows

# Set variables for model
data <- AOM_filtered
response <- data$Mean_rate
predictors <- data %>% select(-Mean_rate)


# Perform a 5-fold cross-validation to find the optimal number of trees (recommended for small datasets). Recommended method by Ridgeway 2024 (author of the gbm package).

#Steps
#The dataset is split into k folds.

cv_folds <- 5

set.seed(123)  # Ensure reproducibility of the model training

gbm_model_cv <- gbm(
  formula = Mean_rate ~ .,
  data = data,
  distribution = "gaussian", #for numerical data 
  n.trees = 10000, #max x value on plot
  interaction.depth = 3, #Elith 2008 recommends values of 2 or 3 for small sample sizes (<250)
  shrinkage = 0.003, #what papers recommend for small sample sizes (Elith 2008; Ridgeway 2024), increased here from 0.001 for better model fit
  #smaller values increase model performance 
  n.minobsinnode = 10, 
  cv.folds = cv_folds, #5-fold cv
  verbose = TRUE
)

# get MSE and compute RMSE. 
#RSME = difference between values predicted by the model and the actual values. "RMSE is a measure of the model's predictive performance, reported here as the average prediction error during cross-validation"

sqrt(min(gbm_model_cv$cv.error))
## 0.27

#Plot shows the cross-validation performance of the boosted regression tree model, showcasing the mean squared error on both training (black) and validation (green) datasets over boosting iterations. The plot helps identify the optimal number of trees to prevent overfitting while maximizing predictive accuracy.


set.seed(123)  
best_ntrees <- gbm.perf(gbm_model_cv, method = "cv", plot.it = TRUE) #here the best # of trees is 5625
print(best_ntrees)

# compute R-squared value
# R2 is a measure of the model's explanatory power, computed here using the predictions from the cross-validation folds. It indicates how well the model is explaining the variance in the target variable within the cross-validation process.

cv_predictions <- predict(gbm_model_cv, data, n.trees = 5625)

# Calculate Total Sum of Squares (TSS)
tss <- sum((data$Mean_rate - mean(data$Mean_rate))^2)

# Calculate Residual Sum of Squares (RSS)
rss <- sum((data$Mean_rate - cv_predictions)^2)

# Calculate R-squared
r_squared <- 1 - (rss / tss)
print(paste("R-squared: ", r_squared))
#0.75

#let's see how the RMSE compares to the standard deviation of the dataset. It should be lower, which indicates the model is demonstrating some ability to learn.

sd(AOM_filtered$Mean_rate)
# 0.31


# Train the final model using the optimal number of trees and all data
final_model <- gbm(
  formula = Mean_rate ~ .,
  data = data,
  distribution = "gaussian",
  n.trees = 5625,
  interaction.depth = 3,
  shrinkage = 0.003,
  n.minobsinnode = 10,
  verbose = TRUE
)


# Summary of the final model
print(summary(final_model))

# Calculate AIC
n <- nrow(data)  # Number of observations
k <- length(final_model$var.levels) + 1  # Number of parameters (number of predictors + 1 for the intercept)
log_likelihood <- -n/2 * log(2 * pi) - n/2 * log(rss/n) - n/2  # Gaussian log-likelihood
aic <- 2 * k - 2 * log_likelihood  # AIC formula
print(paste("AIC: ", aic))

#-73.63

#Variable importance

##1.) Relative Importance

#Each feature in the dataset is assigned a relative importance score. These scores (on the y-axis) are measures of how much each feature contributes to the model's predictive accuracy, normalized so their sum equals 100. The higher the score, the more important the feature is in predicting the target variable. E.g. Dominant Vegetation is the most influential predictor, contributing __% to the model's accuracy.

#"At each split in each tree, gbm computes the improvement in the split-criterion (MSE for regression). gbm then averages the improvement made by each variable across all the trees that the variable is used. The variables with the largest average decrease in MSE are considered most important.


# Extract feature importance
importance <- summary(final_model, plotit = FALSE)

importance_df <- data.frame(
  Feature = importance$var,
  Importance = importance$rel.inf
)

importance_df <- importance_df[order(-importance_df$Importance),]

# Plot feature importance
ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Feature Importance: AOM",
       x = "Features",
       y = "Relative Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#FINAL FIG: Rainbow
rainbowAOM=ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Anaerobic methane oxidation",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Incubation temperature", "Dominant vegetation", "Subpathway", "Salinity", "Depth","Incubation length","Wetland type","Method")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#377eb8",
                               "Subpathway" = "#984ea3", 
                               "Method"="#f781bf",
                               "Wetland_type"="#e6ab02",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#ff7f00")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = c("#e41a1c", "#4daf4a", "#984ea3", "#ff7f00", "#a65628", "#377eb8","#e6ab02", "#f781bf" )), 
        legend.position = "none") 

#FINAL FIG: depth, dom veg, temp highlighted
triAOM=ggplot(importance_df, aes(x = reorder(Feature, -Importance), y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Anaerobic methane oxidation",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Incubation temperature", "Dominant vegetation", "Subpathway", "Salinity", "Depth","Incubation length","Wetland type","Method")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Subpathway" = "#B3B3B3", 
                               "Method"="#B3B3B3",
                               "Wetland_type"="#B3B3B3",
                               "Incubation_temperature"="#e41a1c", 
                               "Salinity"="#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = c("#e41a1c", "#4daf4a", "#999999", "#999999", "#a65628", "#999999","#999999", "#999999")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none")

triAOM_sorted=ggplot(importance_df, aes(x = factor(Feature, levels = c("Dominant_vegetation", 
                                                         "Incubation_temperature", 
                                                         "Depth", "Incubation_length", 
                                                         "Subpathway", "Method", 
                                                         "Wetland_type", "Salinity")), 
                          y = Importance, fill = Feature)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Anaerobic methane oxidation",
       x = "Features",
       y = "Relative Importance") +
  scale_x_discrete(labels = c("Dominant vegetation", "Incubation temperature", "Depth", 
                              "Incubation length", "Subpathway", "Method", 
                              "Wetland type", "Salinity")) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                               "Dominant_vegetation" = "#4daf4a", 
                               "Incubation_length" = "#B3B3B3",
                               "Subpathway" = "#B3B3B3", 
                               "Method" = "#B3B3B3",
                               "Wetland_type" = "#B3B3B3",
                               "Incubation_temperature" = "#e41a1c", 
                               "Salinity" = "#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, 
                                   color = c("#4daf4a", "#e41a1c", "#a65628", 
                                             "#B3B3B3", "#B3B3B3", "#B3B3B3", 
                                             "#B3B3B3", "#B3B3B3")), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.position = "none")

# Normalize importance scores
total_importance <- sum(importance_df$Importance)
importance_df$Normalized_Importance <- importance_df$Importance / total_importance

# Plot the normalized importance scores
ggplot(importance_df, aes(x = reorder(Feature, -Normalized_Importance), y = Normalized_Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Feature Importance: AMO",
       x = NULL,
       y = "Relative Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

##2.) Permutation Importance

# computes the Mean Squared Error (MSE) of the model predictions on the original dataset to those of the permuted dataset.


#Permutation importance scores are calculated by evaluating the change in model performance when each feature's values are randomly shuffled. Initially, the model's performance (Mean Squared Error, MSE) is measured on the original data. Then, for each feature, the values are permuted, and the model's performance is re-evaluated on this permuted data. The importance score for a feature is the difference between the permuted performance and the original performance. A larger positive difference indicates higher importance, as it shows that permuting the feature significantly degrades model performance.
#Features with higher values are crucial for predictions; permuting them significantly increases error (RSME).

# Initial model performance evaluation
original_perf <- with(AOM_filtered, mean((Mean_rate - predict(final_model, newdata = AOM_filtered, n.trees = 5625))^2))

# Storage for importance scores
feature_importance <- setNames(numeric(ncol(AOM_filtered) - 1), names(AOM_filtered)[-which(names(AOM_filtered) == "Mean_rate")])

# Calculate importance for each feature
set.seed(123)  # For reproducibility
for (feature in names(feature_importance)) {
  # Copy the original data
  data_permuted <- AOM_filtered

  # Permute the feature column
  data_permuted[[feature]] <- sample(data_permuted[[feature]])

  # Calculate performance with permuted data
  permuted_perf <- with(data_permuted, mean((Mean_rate - predict(final_model, newdata = data_permuted, n.trees = 5625))^2))

  # Importance score is the loss in performance
  feature_importance[feature] <- permuted_perf - original_perf
}

# Sorting features by importance
feature_importance <- sort(feature_importance, decreasing = TRUE)

# Print the importance
print(feature_importance)

# Convert to data frame for ggplot
importance_df <- data.frame(Feature = names(feature_importance), Importance = feature_importance)

# Plot
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Permutation Importance: AOM", x = NULL, y = "Decrease in Performance (MSE)") +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))


# permutation importance plot with color
permAOM=ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Feature)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Anaerobic methane oxidation",
       x = NULL,
       y = "Decrease in Performance (MSE)") +
  scale_x_discrete(labels = c("Method","Wetland type","Depth", "Incubation length","Subpathway","Salinity", "Dominant vegetation", "Incubation temperature" )) +
  scale_fill_manual(values = c("Depth" = "#a65628", 
                                "Dominant_vegetation" = "#4daf4a", 
                                "Incubation_length" = "#B3B3B3",
                                "Subpathway" = "#B3B3B3", 
                                "Method" = "#B3B3B3",
                                "Wetland_type" = "#B3B3B3",
                                "Incubation_temperature" = "#e41a1c", 
                                "Salinity" = "#B3B3B3")) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1, 
                                   color = c( "#999999","#999999","#a65628", "#999999","#999999","#999999", "#4daf4a","#e41a1c")), 
        plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")


## Partial dependency plots. 

#They visualize the marginal effect of each predictor variable on the predicted outcome, averaging out the effects of other predictors. The y axis represents how the Mean_rate is changing with the change in the given predictor variable.


library(pdp)
library(ggplot2)

for (predictor in setdiff(colnames(AOM_filtered), "Mean_rate")) {
  partial_plot <- partial(final_model, pred.var = predictor, n.trees = 5625)
  
  # Convert partial_plot to data frame
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Determine the type of predictor
  predictor_type <- ifelse(is.numeric(AOM_filtered[[predictor]]), "continuous", "categorical")
  
  # Filter out "Not Reported" data points for categorical predictors
  if (predictor_type == "categorical") {
    partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  }
  
  # Create ggplot based on predictor type
  if (predictor_type == "continuous") {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_line() +
      geom_rug(data = AOM_filtered, aes(x = !!sym(predictor), y = NULL), sides = "t", color = "blue", size = 0.3) +
      labs(title = paste("Anaerobic methane oxidation"),
           x = predictor, y = "Predicted rate") +
      theme_minimal()
  } else {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_point(size = 15, shape = 95) +  
      labs(title = paste("Anaerobic methane oxidation"),
           x = predictor, y = "Predicted rate") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  }
  
  # Print the plot
  print(p)
}

#save individual pdp's (dom_veg, depth, inc_temp) for the paper to be plotted 

library(pdp)
library(ggplot2)

# Generate and store partial dependency plot for Depth
depth_AOM <- {
  partial_plot <- partial(final_model, pred.var = "Depth", n.trees = 5625)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")

  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_line() +
    geom_rug(data = AOM_filtered, aes(x = Depth, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Depth (cm)", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal()
}

# Generate and store partial dependency plot for Dominant Vegetation
dom_veg_AOM <- {
  partial_plot <- partial(final_model, pred.var = "Dominant_vegetation", n.trees = 5625)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Filter out "Not Reported" data points
  partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  
  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_point(size = 8, shape = 95) +  
    labs(title = "Anaerobic methane oxidation",
         x = "Dominant Vegetation", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))
}

# Generate and store partial dependency plot for Dominant Vegetation. Sort x axis categories from lowest to highest predicted rate
dom_veg_AOM_sorted <- {
  partial_plot <- partial(final_model, pred.var = "Dominant_vegetation", n.trees = 5625)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Filter out "Not Reported" data points
  partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  
  ggplot(partial_df, aes(x = reorder(Value, Response), y = Response)) +  # Reorder based on Response
    geom_point(size = 8, shape = 95) +  
    labs(title = "Anaerobic methane oxidation",
         x = "Dominant Vegetation", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))
}


# Generate and store partial dependency plot for Incubation Temperature
incub_temp_AOM <- {
  partial_plot <- partial(final_model, pred.var = "Incubation_temperature", n.trees = 5625)
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")

  ggplot(partial_df, aes(x = Value, y = Response)) +
    geom_line() +
    geom_rug(data = AOM_filtered, aes(x = Incubation_temperature, y = NULL), sides = "t", color = "blue", size = 0.4) +
    labs(
         x = "Incubation Temperature (\u00b0C)", y = expression(paste("Predicted Rate (", mu, "mol CH"[4], "/g/day)"))) +
    theme_minimal()
}

```



Univariate regression tree
```{r}

#packages for regression trees
library(rpart)
library(rpart.plot)

set.seed(123)
tree_univariate_AOM <- rpart(Mean_rate ~. , data = AOM_filtered, method = "anova")

#visualize the tree
rpart.plot(tree_univariate_AOM)

#let's check the residuals
plot(predict(tree_univariate_AOM),residuals(tree_univariate_AOM))

## Pruning the tree

#"Our goal here is to see if a smaller subtree can give us comparable results to the fully grown tree. If yes, we should go for the simpler tree because it reduces the likelihood of overfitting.

#One possible robust strategy of pruning the tree (or stopping the tree to grow) consists of avoiding splitting a partition if the split does not significantly improves the overall quality of the model.

#In rpart package, this is controlled by the complexity parameter (cp), which imposes a penalty to the tree for having two many splits. The default value is 0.01. The higher the cp, the smaller the tree.

#A too small value of cp leads to overfitting and a too large cp value will result to a too small tree. Both cases decrease the predictive performance of the model.

#An optimal cp value can be estimated by testing different cp values and using cross-validation approaches to determine the corresponding prediction accuracy of the model. The best cp is then defined as the one that maximizes the cross-validation accuracy.

# In plotcp() "A good choice of cp for pruning is often the leftmost value for which the mean lies below the horizontal line"
# Here, this is cp = 0.057

plotcp(tree_univariate_AOM)

#Prints a table of optimal prunings based on a complexity parameter
printcp(tree_univariate_AOM)

#visualize more levels
tree_univariate_AOM2 <- rpart(Mean_rate ~. , data = AOM_filtered, method = "anova", control=list(cp=0,xval=10))

plotcp(tree_univariate_AOM2)

printcp(tree_univariate_AOM2)

# Prune the model based on the optimal cp value
tree_uni_AOM_pruned <- prune(tree_univariate_AOM, cp = 0.012)

rpart.plot(tree_uni_AOM_pruned)

printcp(tree_uni_AOM_pruned)


#let's check the residuals
plot(predict(tree_uni_AOM_pruned),residuals(tree_uni_AOM_pruned))


#Extract importance values from tree
importance = tree_uni_AOM_pruned$variable.importance

importance_df = data.frame(
  Feature = names(importance),
  Importance = importance
)

importance_df = importance_df[order(-importance_df$Importance),]

#Gini importance 
ggplot(importance_df, aes(x=reorder(Feature, -Importance), y=Importance))+
  geom_col()+
  theme_minimal()+
  labs(title="Feature Importance: AOM ",
       x="Features", 
       y= "Gini Importance")+
  theme(axis.text.x = element_text(angle = 45, hjust =1))

## Calculate the sum of importance scores
total_importance <- sum(importance_df$Importance)

# Normalize importance scores
importance_df$Normalized_Importance <- importance_df$Importance / total_importance

# Plot the normalized importance scores
ggplot(importance_df, aes(x=reorder(Feature, -Normalized_Importance), y=Normalized_Importance)) +
  geom_col() +
  theme_minimal() +
  labs(title="Relative Feature Importance: AOM",
       x = NULL, 
       y= "Gini Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust =1), )


#Permutation importance 

# Initial model performance evaluation
original_perf <- with(AOM_filtered, mean((Mean_rate - predict(tree_uni_AOM_pruned, newdata = AOM_filtered))^2))

# Storage for importance scores
feature_importance <- setNames(numeric(ncol(AOM_filtered) - 1), names(AOM_filtered)[-which(names(AOM_filtered) == "Mean_rate")])

# Calculate importance for each feature
set.seed(123)  # For reproducibility
for (feature in names(feature_importance)) {
  # Copy the original data
  data_permuted <- AOM_filtered
 
  # Permute the feature column
  data_permuted[[feature]] <- sample(data_permuted[[feature]])
 
  # Calculate performance with permuted data
  permuted_perf <- with(data_permuted, mean((Mean_rate - predict(tree_uni_AOM_pruned, newdata = data_permuted))^2))
 
  # Importance score is the loss in performance
  feature_importance[feature] <- permuted_perf - original_perf
}

# Sorting features by importance
feature_importance <- sort(feature_importance, decreasing = TRUE)

# Print the importance
print(feature_importance)

# Convert to data frame for ggplot
importance_df <- data.frame(Feature = names(feature_importance), Importance = feature_importance)

# Plot
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Permutation Importance: AOM", x = NULL, y = "Decrease in Performance (MSE)") +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))


##Partial dependency plots
library(pdp)
library(ggplot2)

for (predictor in setdiff(colnames(AOM_filtered), "Mean_rate")) {
  partial_plot <- partial(tree_uni_AOM_pruned, pred.var = predictor)
  
  # Convert partial_plot to data frame
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Determine the type of predictor
  predictor_type <- ifelse(is.numeric(AOM_filtered[[predictor]]), "continuous", "categorical")
  
  # Filter out "Not Reported" data points for categorical predictors. We don't want to plot these.
  if (predictor_type == "categorical") {
    partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  }
  
  # Create ggplot based on predictor type
  if (predictor_type == "continuous") {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_line() +
      geom_point() +  
      labs(title = paste("Partial Dependency Plot for", predictor),
           x = predictor, y = "Predicted rate") +
      theme_minimal()
  } else {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_point(size = 15, shape = 95) +  
      labs(title = paste("Partial Dependency Plot for", predictor),
           x = predictor, y = "Predicted rate") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  }
  
  # Print the plot
  print(p)
}


library(pdp)
library(ggplot2)

# Initialize variables to store specific plots
depth_MGEN <- NULL
dom_veg_MGEN <- NULL
incub_temp_MGEN <- NULL

for (predictor in setdiff(colnames(MGEN_filtered), "Mean_rate")) {
  partial_plot <- partial(final_model, pred.var = predictor, n.trees = 5440)
  
  # Convert partial_plot to data frame
  partial_df <- as.data.frame(partial_plot)
  colnames(partial_df) <- c("Value", "Response")
  
  # Determine the type of predictor
  predictor_type <- ifelse(is.numeric(MGEN_filtered[[predictor]]), "continuous", "categorical")
  
  # Filter out "Not Reported" data points for categorical predictors
  if (predictor_type == "categorical") {
    partial_df <- partial_df[partial_df$Value != "Not Reported", ]
  }
  
  # Create ggplot based on predictor type
  if (predictor_type == "continuous") {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_line() +
      geom_point() +  # Add points to the line plot
      geom_rug(data = MGEN_filtered, aes(x = !!sym(predictor), y = NULL), sides = "t", color = "blue", size = 0.3) +
      labs(title = "Methanogenesis", x = predictor, y = "Predicted rate") +
      theme_minimal()
  } else {
    p <- ggplot(partial_df, aes(x = Value, y = Response)) +
      geom_point(size = 15, shape = 95) +  
      labs(title = "Methanogenesis", x = predictor, y = "Predicted rate") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  }
  
  # Save specific plots to variables
  if (predictor == "incubation_temperature") {
    incub_temp_MGEN <- p
  } else if (predictor == "dominant_vegetation") {
    dom_veg_MGEN <- p
  } else if (predictor == "depth") {
    depth_MGEN <- p
  }
  
  # Print the plot
  print(p)
}

```

Shapley values from pruned univariate regression tree
```{r}

# Load necessary libraries
library(kernelshap)
library(shapviz)
library(ggplot2)


# Set seed for reproducibility
set.seed(123)

# Define the full dataset for both rows to be explained and background data. I'm doing this since my dataset is so small (n=93).

X <- AOM_filtered[, -which(names(AOM_filtered) == "Mean_rate")]  # Exclude target variable
bg_X <- X  # Use the same dataset for background data

# Define a prediction function for the model
pred_fn <- function(model, newdata) {
  predict(model, newdata = newdata)
}

# Calculate SHAP values using kernelshap
shap_values <- kernelshap(tree_uni_AOM_pruned, X = X, bg_X = bg_X, pred_wrapper = pred_fn, nsim = 100)

# Convert SHAP values to a shapviz object for analysis
shap_viz <- shapviz(shap_values)

# Analyze and visualize SHAP values

# Plot feature importance
sv_importance(shap_viz) #looks similar to permutation importance plot
sv_importance(shap_viz, kind = "bee") # plot that shows directionality is not well suited to my dataset bc the most important predictors (dominant vegetation, wetland type) are categorical.  

```

Graph final plots together
```{r}
library(ggpubr)

##feature importance

#rainbow plots
ggarrange(rainbowMGEN, rainbowMOx, rainbowAOM, ncol = 3, nrow = 1)

#tricolor plots
ggarrange(triMGEN, triMOx, triAOM, ncol = 3, nrow = 1)

#tricolor plots, x axis sorted, vertical position 
ggarrange(triMGEN_sorted, triMOx_sorted, triAOM_sorted, ncol = 1, nrow = 3)

##partial dependency plots vertically grouped
ggarrange(dom_veg_MGEN,incub_temp_MGEN, depth_MGEN, dom_veg_MOx, incub_temp_MOx, depth_MOx, dom_veg_AOM, incub_temp_AOM, depth_AOM, ncol = 3, nrow = 3)

##partial dependency plots horizontally grouped
ggarrange(dom_veg_MGEN,dom_veg_MOx, dom_veg_AOM, incub_temp_MGEN, incub_temp_MOx, incub_temp_AOM,depth_MGEN, depth_MOx, depth_AOM, ncol = 3, nrow = 3)

##partial dependency plots horizontally grouped, dom veg sorted from lowest to highest y value
ggarrange(dom_veg_MGEN_sorted,dom_veg_MOx_sorted, dom_veg_AOM_sorted, incub_temp_MGEN, incub_temp_MOx, incub_temp_AOM,depth_MGEN, depth_MOx, depth_AOM, ncol = 3, nrow = 3)

#permutation plots
ggarrange(permMGEN,permMOx, permAOM, ncol = 3, nrow = 1)
```

